Single-cell profiling reveals the intratumor heterogeneity and immunosuppressive microenvironment in cervical adenocarcinoma

Version 1

Abstract

Abstract

Background

Cervical adenocarcinoma (ADC) is more aggressive compared to other types of cervical cancer (CC), such as squamous cell carcinoma (SCC). The tumor immune microenvironment (TIME) and tumor heterogeneity are recognized as pivotal factors in cancer progression and therapy. However, the disparities in TIME and heterogeneity between ADC and SCC are poorly understood.

Methods

We performed single-cell RNA sequencing on 11 samples of ADC tumor tissues, with other 4 SCC samples served as controls. The immunochemistry and multiplexed immunofluorescence were conducted to validate our findings.

Results

Compared to SCC, ADC exhibited unique enrichments in several sub-clusters of epithelial cells with elevated stemness and hyper-malignant features, including the Epi_10_CYSTM1 cluster. ADC displayed a highly immunosuppressive environment characterized by the enrichment of regulatory T cells (Tregs) and tumor-promoting neutrophils. The Epi_10_CYSTM1 cluster recruits Tregs via ALCAM-CD6 signaling, while Tregs reciprocally induce stemness in the Epi_10_CYSTM1 cluster through TGFβ signaling. Importantly, our study revealed that the Epi_10_CYSTM1 cluster could serve as a valuable predictor of lymph node metastasis for CC patients.

Conclusions

This study highlights the significance of ADC-specific cell clusters in establishing a highly immunosuppressive microenvironment, ultimately contributing to the heightened aggressiveness and poorer prognosis of ADC compared to SCC.

Introduction

Cervical cancer (CC) is the fourth leading cause of cancer-related death among all types of women malignancies (Sung et al., 2021). The two major histological types of CC are squamous cell carcinoma of cervix (SCC) and adenocarcinoma of cervix (ADC), accounting for 70% and 25% of all cases, respectively (Cohen, Jhingran, Oaknin, & Denny, 2019). ADC patients tend to have a higher rate of rapid progression, relapse and insensitivity to chemotherapy, radiotherapy and even immunotherapy, making it a more aggressive phenotype than SCC (Sasieni, Castanon, & Cuzick, 2009). Unfortunately, the reasons are not yet well understood due to its rarity. Furthermore, the standard treatment for both types of CC is not stratified based on pathologic types, which can result in treatment failures of ADC patients. Therefore, there is an urgent need for a comprehensive understanding of ADC.

In 2022, the National Comprehensive Cancer Network (NCCN) guidelines recommended pembrolizumab (Keytruda ®, a PD1 blockade) as the first-line combination therapy for late-stage, recurrent and metastatic CC patients with positive PD-L1 expressions (Colombo et al., 2021). It is a milestone that immunotherapy has been updated as a first-line treatment for CC. While it has been reported that 30%-50% of CC patients exhibit positive PD-L1 expression, however, only an estimated 20%-30% are expected to respond positively to PD-1/PD-L1 blockade therapy (Omenai, Ajani, & Okolo, 2022; Sun et al., 2020). Therefore, CC is characterized as an immunosuppressive cancer type, due to modulation of the tumor immune microenvironment (TIME) (Cao et al., 2023). TIME, consisting of tumor cells, immune cells, cytokines and their interactions, predominantly determines the efficacy of immunotherapy (Lv et al., 2022). The mechanisms of tumor immune evasion within the TIME are similar across different types of cancer, involving the deactivation of immune surveillance pathways or recruitment of immunorepressive cells (Vinay et al., 2015). However, the specific characteristics of the TIME in ADC, which is known as insensitive to immunotherapy, remain poorly understood.

Persistent infection with human papillomavirus (HPV) is the primary cause of CC (Mann, Singh, & Kumar, 2023). Approximately 85% of CC patients are identified with HPV infection, while 5%-15% of patients test negative for HPV (X. Yang et al., 2023). It is reported that HPV-negative patients tend to experience worse clinical outcomes, including rapid metastasis to distant organs, and an increased susceptibility to developing resistance to immunotherapy (Fernandes, Viveros-Carreño, Hoegl, Ávila, & Pareja, 2022). However, current therapeutic strategies for CC do not take into account the HPV infection status as a stratification factor. Therefore, it is imperative to investigate the precise mechanisms underlying the development of HPV-negatively infected tumors. In addition, an important and unavoidable clinical issue in CC treatment is post-surgical upstaging, which affects the disease outcome and treatment approach.

To extensively investigate the cellular and molecular heterogeneity of CC, particularly ADC, the utilization of single-cell RNA sequencing (scRNA-seq) technique is crucially needed. In the current study, we focused on the remodeling of TIME in ADC and performed scRNA-seq analysis on 11 ADC samples, with another 4 SCC samples included as controls. By high-throughput analysis, we are able to characterize the landscape of TIME in ADC, with different types of tumor-promoting cells being recruited in the vicinity of the tumor, thus creating an immunosuppressive environment in ADC. The cellular crosstalk analysis provides valuable insights into the interactions between different cell types within the TIME. Furthermore, we have conducted a screening of novel signature genes as prognostic biomarkers for ADC. These biomarkers have the potential to accurately predict lymph node metastasis and may guide personalized treatment decisions more effectively.

Methods

Clinical sample collection

The 15 cases of fresh cervical cancer tissues were collected from Hunan Cancer Hospital, with informed consent obtained from 15 independent patients. The inclusion criteria are as follows: 1) The diagnosis of CC, with pathological type of ADC or SCC, should be confirmed by pathologists; 2) Pre-surgical FIGO stages should be within the range from IB (which means the tumor mass should be detectable via CT or MRI) to IIA (which means no parametrial involvement, nor distant metastasis), in which patients have explicit indication for radical surgery following NCCN guidelines; 3) No neo-adjuvant treatments should be given before the surgery, including chemotherapy and radiotherapy; 4) The surgery options should be type-C radical hysterectomy and include pelvic lymphadenectomy with/ without para-aortic lymphadenectomy. Detailed information of the 15 patients was listed in Supplementary Table 1. This study was supervised and approved by the Ethics Committee Board of Hunan Cancer Hospital.

Sample preparation for single-cell isolation

Immediately after surgical removal, the fresh tumor tissues were washed by saline for 3 times and stored in cold storage solution (MACS® Tissue Storage Solution) for transportation. To prepare high-quality samples, the tissues were cut into pieces to 1mm3 and were incubated at 37°C for 30 minutes. 0.25% trypsin solution was used to digest the tissue pieces at 37°C for 10 minutes. Then single-cell samples were filtered through a 70-μm meshed filter and suspended in a red blood cell lysis buffer (MACS® Red Blood Cell Lysis Solution) for dissolution. Before instrumental sequencing, suspension of single cells were visualizing by TC20TM Automated Cell Counter to assess the cell viability. Each sample was confirmed to have a cell viability rate of over 80%.

Single-cell RNA sequencing

The suspension of single cells was transferred onto the 10X Chromium Single-Cell instrument to generate single-cell beads in the emulsion (GEMs). Then the scRNA-seq libraries were constructed by using the Chromium Next GEM Single Cell 3’ Reagent Kits v3 (10X Genomics, USA) and sequenced by using the sequencer Novaseq6000 (Illumina, USA). All procedures were following the manufacturer’s protocol.

Dimension reduction and cell clustering

The gene expression matrixes were generated via the Cell Ranger toolkit (version 5.0.0) of 10X Genomics platform. The package of Seurat (version 4.0.5 R) was used to normalize, convert and format data. With the function of RunPCA, the principal component analysis (PCA) was used for data dimension reduction. The functions of FindNeighbors and FindClusters were used for cell clustering. For visualization, the RunUMAP function was performed. In each unsupervised cell cluster, the gene markers were identified by the function of FindAllMarkers with comparison to other clusters. For data processing, cells with low qualities, which showed <200 expressed genes or >25% mitochondrial UMIs (unique molecular identifiers), were excluded.

Cell type annotation

The cell typing was conducted and annotated according to the selected gene markers via CellMarker database and publications (C. Li et al., 2022; Qiu et al., 2023; Xue et al., 2022). The expression of differential expressed genes (DEGs) was used to determine each cell type following the rules: Log2Foldchange should be > 0.2 and adjusted-P values should be < 0.05.

Pseudotime trajectory analysis

To identify the developmental changes of epithelial cells, the pseudotime trajectory analyses were performed by using Monocle2 R package (version 2.22.0), in which Seurat was used as input. Calculated by the Monocle algorithm, the top 50-1000 genes were conducted for differential GeneTest. Then, the DDR-Tree and default parameters were generated to visualize the translational relationship and developmental orders among different epithelial sub-clusters.

Cell malignancy scoring

Among these epithelial cells, the malignant score of each single cell was predicted by using scCancer R package (Guo et al., 2021). The malignant cells were distinguished from non-malignant cells by inferring large-scale copy-number variations (CNVs), which was calculated by the interCNV R package as described (Guo et al., 2021; Xue et al., 2022).

Signaling pathway enrichment analysis

The biological signaling pathway of each cell cluster was identified by performing Gene Ontology (GO) analyses and Hallmark Pathway enrichment analyses, according to the Molecular Signature Database (MSigDB version 7.5.1). Then DEGs of each cell cluster were converted to ClusterProfiler (version 4.2.2) package for analysis of predicted functions.

CytoTRACE

In order to predict the differential state of cells, scRNA-seq derived data were pass to the CytoTRACE software (version 0.3.3) with R package for further analyses.

Survival analysis

The CESC dataset from TCGA database was downloaded from Xena, with clinical characteristics, survival endpoints and treatments were included. The Kaplan-Meier survival curves were generated by using R software.

Cellular communication analysis

The possible interactions, with ligand-to-receptor communicating signaling pathway, between two types of cells were analyzed using the R packages of CellChat (version 1.5.0). Seurat was used for normalization and the rank with significance was calculated. The probability of ligand-to-receptor interaction in each signaling pathway was plotted by P-value and intensity.

Immunohistochemistry (IHC)

The paraffin-embedded ADC samples (whether post-surgical or biopsy samples) were sectioned into 4μm slides. The protocol of IHC strictly followed our previous published works (Peng et al., 2019). The primary antibody of SLC26A3 was purchased from Santa Cruz Biotechnology (Cat# sc-376187) and diluted at the ratio of 1:20. The protein expression of SLC26A3 was determined using the method of H-scoring: H-score = intensity score × percentage of positivity. The intensity was scored as: Negative=0, Weak=1, Moderate=2, Strong=3. Samples were grouped as the higher expression group (H-score ≥ 150) and the lower expression (H-score < 150) group of SLC26A3.

Multiplexed immunofluorescence (multi-IF)

The multi-IF was performed using the Opal 6-Plex Manual Detection Kit (NEL811001KT, Akoya), following the manufacturer’s instructions. Generally, the slides were stained with each primary antibody sequentially following the staining cycle as: (1) antigen retrieval by citrate (pH=6) at water bath for 30 minutes; (2) independent primary antibody incubation at 4°C atmosphere overnight or 37°C atmosphere for 1 hour; (3) common HRP-crosslinked secondary antibody incubation at 37°C atmosphere for 30 minutes; (4) independent opal reactive fluorophore solution at intended wavelength (opal 570 was excluded in order to save for FOXP3 signal). Each independent primary antibody started a new cycle. After cycles finished, the slides were prepared for incubation of FOXP3 monoclonal antibody with eFluor™ 570 (Thermo Fisher, Cat#41-4777-82) conjugated, as well as DAPI (Sigma) for nucleus staining.

Statistics

Specific analyses of scRNA-seq data were decoded and visualized by using the R software. Statistics were processed using IBM-SPSS Statistics 20.0 software and R software as well, including Student’s t-test, two-sided Wilcoxon test, etc. Correlations analysis was performed by the Chi-square test. Survival analysis was performed by the Kaplan-Meier method. P values < 0.05 were considered as statistically significant.

Results

Single-cell transcriptomic landscape of cervical cancer

We collected 15 samples of cervical cancer tissues from 15 individual patients immediately after undergoing radical hysterectomy surgery. Among these cases, 11 were ADC types, with 5 being HPV-negative and 6 being HPV-positive. The remaining 4 cases were SCC types, with an even distribution of HPV-positive and HPV-negative status (Supplementary Table 1). Prior to surgery, all 15 patients were diagnosed as early FIGO stages (stage I ∼IIA) based on evaluation of CT, MRI, and physical examination. However, after surgery, 5 cases turned out to be upstaged with pathological evidence of lymph node metastasis (FIGO stage IIICP), requiring theses patients to receive radiation therapy according to NCCN guidelines. Therefore, in addition to investigate the difference between histological types and HPV infection status, we also focused on the issue of post-surgical upstaging by comparing different FIGO stages.

In order to analyze the genomic profiling, we resolved these samples at the single-cell level using the 10X genomics chromium platform. The cell viability and sample quality were ensured beforehand, and a total of 102,012 qualified cells were utilized for further analysis (Fig. 1A). The technique of Uniform Manifold Approximation and Projection (UMAP) was utilized to reduce the dimension of gene profiling data for visualization. The obtained cells were primarily categorized into 12 cell clusters, which were further re-clustered into 9 distinct cell types, based on specific gene markers (C. Li et al., 2022) (Fig. 1B-C). These cell types included T cells (38,627 cells in total, proportioning 37.87%, marked with PTPRC, CD3D, CD3E and CD3G), epithelial cells (32,199 cells in total, proportioning 31.56%, marked with EPCAM, SLP1 and CD24), neutrophils (12,586 cells in total, proportioning 12.34%, marked with CSF3R), macrophages (6,798 cells in total, proportioning 12.34%, marked with CD68 and CD163), fibroblasts (4,258 cells in total, proportioning 4.17%, marked with COL1A2 and DCN), plasma cells (4,044 cells in total, proportioning 3.96%, marked with JCHAIN), endothelial cells (1,389 cells in total, proportioning 1.36%, marked with ENG and VWF), B cells (1,326 cells in total, proportioning 1.30%, marked with CD79A and MS4A1) and mast cells (785 cells in total, proportioning 0.77%, marked with MS4A2) (Fig. 1D-E and Supplementary Table 2).

The single-cell genomic atlas of cervical cancer.

A. The schematic design of sample collection, single-cell RNA sequencing, data processing and clinical validating through 10X genomics platform. The sketch of scRNA-sq machine, as well as the plot of dimension-reduction visualization, was originally cited from the official website of 10X genonics: https://www.10xgenomics.com. B. UMAP plots of all single cells by original clustering and cell type re-clustering, according to the published lineage-specific marker genes, respectively. C. Violin plots demonstrating the expression of marker genes that correspond to each of the major cell types. D. UMAP plots demonstrating the most typical marker gene specific to each type of cell cluster. E. UMAP plot and histogram plot showing the occupation ratios of the major cell types in each individual sample, with HPV status and histological type summarized. F. UMAP (up) and histogram (down) plots to show the differences of distribution and proportion of each cell type between different histological types (ADC vs. SCC).

According to our scRNA-seq data, ADC exhibited a significant higher proportion of neutrophils than SCC (20.05% vs. 4.23%). The proportion of T cells was also higher in ADC (41.13%) compared to SCC (35.48%). Conversely, SCC showed a nearly threefold enrichment of epithelial cell (47.43%) compared to ADC (16.48%) (Fig. 1F). When considering the HPV infection status, HPV-negative patients demonstrated decreased proportions of epithelial cells (7.78% vs. 28.38%) and neutrophils (16.66% vs. 24.68%) compared to HPV-positive patients. The proportion of T cells enriched in HPV-negative cases (45.77%) was higher than that in HPV-positive cases (32.42%). HPV-negative patients exhibited prominently higher proportions of plasma cells (10.13% vs. 1.99%) and B cells (2.84% vs. 0.77%) compared to HPV-positive patients, even though their proportions were lower than T cells, epithelial cells and neutrophils (Supplementary Fig. 1A). Furthermore, we conducted a comparison between early-stage and late-stage cases, revealing that late-stage patients have a higher proportion of epithelial cells (35.50%) than early-stage cases (28.82%) (Supplementary Fig. 1B). In summary, these data describes primarily a complex microenvironment in cervical cancers, especially in the ADC type, with varied composition of immune and tumor cells under different conditions.

The sub-clusters of epithelial cells exhibit elevated stem-like features to promote the aggressiveness of ADC

The aforehand data has shown that the majority of the CC tumor is composed of epithelial cells. In order to explore the characteristics of these epithelial cells within the TIME of ADC, we further classified them into 12 sub-clusters based on the subsets of differently expressed genes (DEGs) in each group (Fig. 2A). Each sub-cluster was named based on the most prominently up-regulated gene in each panel of signature genes (Fig. 2B-C). Several epithelial cell clusters were strongly enriched in ADC than SCC, such as Epi_04_TFF2, Epi_06_TMC5, Epi_07_CAPS and Epi_08_SCGB3A1. Notably, Epi_09_SST, Epi_10_CYSTM1, Epi_11_REG1A and Epi_12_RRAD were exclusively identified in ADC, not in SCC (Fig. 2D). The stratified enrichment of different cluster between ADC and SCC has captured our attention for further investigation. Additionally, Epi_09_SST, Epi_10_CYSTM1 and Epi_11_REG1A were solely enriched in HPV-positive cases, although no distinctive clusters had been found in HPV-negative patients (Supplementary Fig. 2A). When comparing samples from patients in the early stages with those who have lymph node metastasis, 3 clusters (Epi_02_IGLC2, Epi_05_CCL5 and Epi_12_RRAD) were slightly increased among the late-stage patients. Surprisingly, we found that the sub-cluster Epi_10_CYSTM1 was exclusively present in stage IIIC patients, making it a potential target for identification of the biomarkers for late stage patients (Supplementary Fig. 2B).

The scRNA-seq data reveals the malignant features of tumor epithelial cells.

A. UMAP plots showing the distribution of epithelial cells in 15 samples and the sub-clustering into 12 clusters according to ectopic gene expressions. Each cluster of epithelial cells was named using the most highly enriched gene. B. Heatmap plot showing the annotation of each epithelial sub-cluster with top 5 DEGs. C. UMAP plots demonstrating the most specific gene as the marker for each sub-cluster. D. UMAP (left) and histogram (right) plots to compare the differences of distribution and proportion of each epithelial cell sub-cluster between different histological types (ADC vs. SCC). E. UMAP plot of CytoTRACE showing the thermal imaging projection of predicted developmental order (left) and histogram plot showing the ranking of CytoTRACE scores (right). F. UMAP plot of malignancy analysis demonstrating the cell malignancy features (up) and ranking of scores (down). G. GOBP analyses showing the top 15 enriched signaling pathways of Epi_10_CYSTM1 that are more active in ADC than SCC (up), particularly pathways related to cell-to-cell interactions (down). The histogram data are transformed from -Log10(P-value) for visualization. Abbreviations: Ig: immunoglubolin; MHC: major histocompatibility complex.

We further evaluated the features of each epithelial cluster, especially their roles in regulating tumorigenesis. To assess the differentiation potential of each cell cluster, we employed the Cellular Trajectory Reconstruction Analysis using gene Counts and Expression (CytoTRACE) method (Fig. 2E). The results showed that clusters of Epi_07_CAPS, Epi_09_SST, Epi_10_CYSTM1 and Epi_12_RRAD ranked the top four in CytoTRACE score, indicating a significant potential for pluripotency and differentiation, also referred to stem-like features. Interestingly, Epi_09_SST, Epi_10_CYSTM1 and Epi_12_RRAD were exclusively enriched in ADC. The proportion of cluster Epi_07_CAPS, which acquired the highest level of stemness, was overwhelmingly higher in ADC, although Epi_07_CAPS was also identified in SCC (Fig. 2D). Subsequently, pseudotime analysis was conducted to elucidate the differentiation trajectories. The results showed that Epi_07_CAPS, Epi_09_SST and Epi_10_CYSTM1 were positioned towards the end of pseudotime developmental trajectory, which was in consistence with their high stem-like characteristics and indicated a higher degree of malignancy in ADC than SCC (Supplementary Fig. 2C). In addition, we performed the malignancy scoring analysis, which demonstrated that cluster Epi_10_CYSTM1 exhibited the highest degree of malignancy compared to other clusters (Fig. 2F). Interestingly, cluster Epi_10_CYSTM1 displayed a remarkably diverse developmental profile and was exclusively identified in ADC. Conversely, Epi_07_CAPS and Epi_12_RRAD, which exhibited high stem-like characteristics, were ranking lower in terms of malignancy scoring when compared to other clusters. This observation may partially indicate that high stemness cluster Epi_10_CYSTM1 is essential for ADC to present more aggressive features.

Next, we proceeded to further investigate cluster Epi_10_CYSTM1, particularly in the context of ADC. To this end, Gene Ontology Biological Process (GOBP) analysis on specific signaling pathways based on the gene expression pattern observed in each cell cluster was performed (Fig. 2G and Supplementary Fig. 2D). The results revealed that ADC-specific epithelial cell clusters, particularly the most malignant cluster Epi_10_CYSTM1, tended to be more active in regulating immunity-related pathways, which indicated epithelial sub-cluster might be associated with dysfunction of immunity. Notably, Epi_10_CYSTM1 cluster also played a crucial role in regulating cell-to-cell interaction (Fig. 2G). This gives us a hint that ADC cells exert their tumor-promoting functions via interacting with other types of cells, such as neutrophils or immune cells. Therefore, further investigation is warranted to explore the intricate crosstalk networks between epithelial cells and other cell types.

ADC-specific epithelial cluster-derived gene SLC26A3 is a potential prognostic marker for lymph node metastasis

Given that Epi_10_CYTSM1 represents the most aggressive cell cluster in ADC, we further examined the relevance of this cluster and the clinical features of CC. As an alternative, we also examined Epi_12_RRAD (Supplementary Fig. 3A), which was another malignant cluster predominantly found in ADC. We firstly checked the DEGs of these two clusters. In Epi_12_RRAD cluster, we identified Insulin-like growth factor 2 (IGF2) and Alcohol dehydrogenase 1C (ADH1C), which were exclusively expressed in Epi_12_RRAD, as two most specific genes of this cluster (Supplementary Fig. 3B). However, the immunohistochemistry (IHC) staining showed that IGF2 was almost negative across cases with different clinical stages (Supplementary Fig. 3C). On the other hand, ADH1C was strongly expressed in both early-stage and late-stage cases (Supplementary Fig. 3C). Likewise, the survival analysis of Epi_12_RRAD-enriched gene signatures did not demonstrate a significant difference between the two groups (Supplementary Fig. 3D). These results suggest that the Epi_12_RRAD cluster may not play a significant role in the progression of ADC.

In contrast, the panel genes enriched in Epi_10_CYSTM1 might potentially predict poor prognosis of CC patients based on TCGA database (Fig. 3C). In the cluster of Epi_10_CYSTM1, the DEG with the highest expression is CYSTM1 (Fig. 3A). However, CYSTM1 was also expressed in other cell clusters and showed low specificity (Fig. 3B). On the other hand, SLC26A3, ORM1 and ORM2 were specifically enriched in Epi_10_CYSTM1 and could be considered as potential candidate biomarkers for representation of this cluster. Nevertheless, ORM1/OMR2 were not satisfied to distinguish the severity of CC cases, due to an irrelevance between expression intensity and clinical stages (Supplementary Fig. 3E). Interestingly, the gene of SLC26A3 derived from Epi_10_CYSTM1 showed the most potential of prognostic value for CC patients (Supplementary Fig. 3E).

SLC26A3 is identified as a potential prognostic and diagnostic indicator for lymph node metastasis of CC patients.

A. Heatmap showing the top 50 DEGs of Epi_10_CYSTM1 cluster in comparison to all the other sub-clusters (Red: Epi_10_CYSTM1 cluster; Green: other epithelial cell sub-clusters). B. Violin plots showing the expression difference between two groups (top), with UMAP plots (bottom) demonstrating the specificity of each candidate gene, filtering SLC26A3 as the most identical marker for this sub-cluster. C. Kaplan-Meier curve showing the overall survival rates of CC patients stratified by the top 50 genes-scaled signature of Epi_CYSTM1. D. IHC staining showing the protein expression of SLC26A3 in surgically-resected ADC samples which are classified as early stages (FIGO stage I-IIA) and late stages (FIGO stage IIIC1∼2p). Images from 6 individual cases were shown as representatives for each group. E. IHC staining showing the protein expression of SLC26A3 in biopsy ADC samples which are classified as early stages (FIGO stage I-IIA) and late stages (FIGO stage IIIC1∼2p). Images from 4 individual cases were shown as representatives for each group. The method of H-score is used and the scoring system is as follows: negative (0), weak (1), intermediate (2), and strong (3). Expression is quantified by the H-score method.

An important and unavoidable clinical issue in CC treatment is post-surgical upstaging, which affects the disease outcome and treatment approach. In order to examine the fact of post-surgical upstaging (or misdiagnose of early-stage cases), we merged data from two independent patient cohorts obtained from two clinical centers: Xiangya hospital (Cohort 1) and Hunan cancer hospital (Cohort 2). The criterion for objective selection was outlined below (Supplementary Fig. 3F). The rates of upstaging for each center are 10.85% (23 out of 212 in Cohort 1) and 15.31% (214 out of 1398 in Cohort 2), respectively. These rates fall within the range reported by previous studies (Dabi et al., 2018; Thelissen et al., 2022). Interestingly, both cohorts of data revealed that the misdiagnose rate was significantly higher in HPV-negative patients than HPV-positive patients, especially among ADC patients. From the data of Cohort 2, it was observed that ADC patients were more likely to stage up post-surgically. In Cohort 1, we can also observed the same tendency although without significant difference between different histological types, probably due to a smaller sample size (Table. 1-2).

The identification of Epi_10_CYSTM1 as the only cell cluster found in patients with stage IIICp raises the possibility that this cluster may be a potential marker to diagnose patients with lymph node metastasis. Therefore, we detected SLC26A3 expression on biopsy and post-surgical ADC samples to examine whether it was associated with lymph node metastasis. The results from IHC staining showed that ADC patients with stage IIIC had a higher expression of SLC26A3 (Fig. 3D, E and Table 3, 4). In summary, our results propose that the Epi_10_CYSTM1 cluster could function as a diagnostic marker to predict both poor prognosis and lymph node metastasis in ADC patients. Moreover, SLC26A3 could be employed as a marker for the Epi_10_CYSTM1 cluster, aiding in the diagnosis of lymph node metastasis to prevent post-surgical upstaging in ADC patients in the future.

Enrichment of regulatory T cells indicates an immunosuppressive status of ADC

T cells play a crucial role as the primary effectors of tumor immunity. Our data reveals that T cells constitute the predominant cell population within the TIME of cervical cancer (Fig. 1A). A total of 52,297 T cells were categorized into 8 main clusters (Fig. 4A). The gene expression patterns of T cells in CC were in accord with those observed in other types of cancers (Sorin et al., 2023; Xue et al., 2022). Based on canonical genes (C. Li et al., 2022), these cells were re-clustered as the following 4 types of T cells: exhausted T cells (marked with CD8, TIGIT, PDCD1, LAG3, HAVCR2), cytotoxic T cells (marked with CD8 and CD3), regulatory T cells (Treg, marked with CD4, FOXP3 and IL2RA), activated T cells (marked with CD8, C69 and CD3G) (Fig. 4A-C). The gene subset of cytotoxic T cells exhibited a higher expression of KRT17, KRT19, while Tregs showed a higher expression of genes such as CCL4, GZMB and CCL3 (Fig. 4D). When comparing the checkpoint pathway states among different T cell clusters, we observed relatively higher levels of LAG3 and PDCD1 (PD1) in exhausted T cells, while CTLA4 and TIG1 were more highly expressed in Tregs. These genes were identified to present immunosuppressive functions (Fig. 4E). On the other hand, certain genes, such as NKG7, PRF1, INFG and TNFRSF, which indicate the activation of immune checkpoint pathways, exhibited low expression levels in cytotoxic T cells and activated T cells (Fig. 4E). These findings lead us to hypothesize that the establishment of an immunosuppressive TIME in ADC may involve the recruitment of regulatory T cells (Tregs) to the tumor area and the subsequent inactivation of a substantial proportion of cytotoxic T cells.

Cellular and molecular heterogeneity of T cells in ADC.

A. UMAP plots showing the sample distribution (left), original cell clustering of T cells (middle, 8 clusters in total) and re-clustering of T cell subtypes (right, 4 major sub-clusters: exhausted T cell, cytotoxic T cell, regulatory T cell and activated T cell) according to acknowledged marker genes. B. Violin plot showing the expression of specific marker genes that annotate each sub-type of T cells. C. UMAP plots showing the widely recognized classification markers that denote each type of T cell sub-cluster. D. Dot heatmap plot of row-scaled expression of overexpressed genes for each sub-cluster of T cells. E. Dot heatmap plots that demonstrate the level of marker genes representing the signaling pathways of immune checkpoint activation and inhibition for grouped sub-clusters. F. Differences of distribution and proportion of each T cell sub-cluster between different histological types (ADC vs. SCC). G. GOBP analysis of Treg-enriched immunity-related pathways that are more active in ADC than SCC. The histogram data are transformed from -Log10 (P-value). Abbreviation: Treg: regulatory T cell.

We further compared different status of CC with different attribution of T cell subtypes. It was demonstrated that ADC had a significantly higher proportion of Tregs compared to SCC (Fig. 4F). Meanwhile, ADC patients who tested negative for HPV showed a higher enrichment of Tregs and a lower proportion of activated T cells compared to those with HPV-positive status (Supplementary Fig. 4A). These findings shed light on the underlying mechanism for immunotherapy insensitivity of ADC, especially in HPV-negative status. Furthermore, analyses of GOBP have shown that Tregs were more active in regulating immune activities in ADC than SCC, indicating a greater tendency for Tregs to be activated in the TIME of ADC (Fig. 4G). In ADC, the proportion of activated T cells was higher compared to SCC. However, the GOBP analysis suggested that their leading roles were to promote the apoptosis of immune cells than to exert cytotoxicity (Supplementary Fig. 4D). In addition, in HPV-negative cases, activated T cells demonstrated a stronger capacity to interact with other cell types, including neutrophils and epithelial cells (Supplementary Fig. 4E). Consequently, it is further proved that the enrichment of Tregs in ADC does execute immunosuppressive functions.

Tumor associated neutrophils (TANs) surrounding ADC tumor area contribute to the formation of a malignant microenvironment

The large-scale enrichment of specific gene subsets in neutrophils in CC prompted us to further examine their roles. A total of 12,586 neutrophils were detected and the composition proportion in ADC (20.05%) was almost five times higher than that in SCC (4.23%) (Fig. 1F). Based on this observation, we hypothesize that ADC-specific neutrophil clusters may be related to poor prognosis and increased malignancy. All the cells were initially grouped into 6 clusters based on the gene expression modules (Fig. 5A). In the context of cancer, tumor associated neutrophils (TANs) can exhibit dual functions, either promoting tumor (pro-tumor TANs) or inhibiting tumor progression (anti-tumor TANs). Based on the consensus markers identified by Jaillon et al. (Jaillon et al., 2020) and Xue et al.(Xue et al., 2022), we performed further clustering of the neutrophils in our dataset, resulting in the identification of 3 sub-types of TANs: pro-tumor TANs (Neu_03, marked with CD66, CD11 and MT1X), anti-tumor TANs (Neu_02, marked with CD66, CD11, CCL5, KRT5 and ELF3), TANs with isg (interferon stimulated genes) and other undefined types (Neu_01, distinguished by CD66, CD11, IFITM2, S100A8 and LST1) (Fig. 5B-C). Generally, the Neu_01 sub-cluster, which consists of TANs with isg with the subset of specific genes, accounted for the largest proportion among all neutrophils. Notably, the presence of pro-tumor TANs was significantly higher in ADC compared to SCC (Fig. 5D), as well as in HPV-negative cases compared to HPV-positive cases (Supplementary Fig. 5A). On the other hand, the proportion of anti-tumor TANs in ADC was predominantly lower than that in SCC (Fig. 5D).

The heterogeneity of tumor associated neutrophils in ADC.

A. UMAP plots showing the distribution of neutrophils in 15 samples (left), original clustering (middle) the re-clustering (right) into 3 sub-clusters according to gene markers of tumor-associated neutrophils (TANs). B. Dot heatmap showing the top 5 DEGs in each sub-cluster of TANs. C. UMAP plots for annotation of each sub-type of TANs with published marker genes presented: pro-tumor TANs, anti-tumor TANs and TANs with isg. D. UMAP and histogram plots to compare the differences of distribution and proportion of each sub-types of TANs between different histological types. E. GOBP analyses of signaling pathways that are more active in ADC than SCC, in terms of the pro-tumor TANs cluster. The histogram data are transformed from -Log10 (P-value). F. Kaplan-Meier curve showing the overall survival rate of CC patients stratified by the top 90 genes-scaled signature of pro-TANs.

We further conducted the GOBP analysis on gene subsets derived from pro-tumor TANs to explore their potential functions. The findings revealed that pro-tumor TANs in ADC were more active in regulating cell-to-cell interactions and chemotaxis. Specifically, they were involved in pathways such as chemokine-mediated or cytokine-mediated signaling, neutrophil activation, and receptor internalization (Fig. 5E). These results suggest that pro-tumor TANs may engage in extensive communication with various cell types. Interestingly, pro-tumor TANs exhibited enhanced activation in mediating inhibitory pathways related to immunity (Fig. 5E). These pathways include the negative regulation of leukocyte proliferation and lymphocyte activation, as well as the suppression of IL-10 production, which is known to promote immune responses (Saraiva, Vieira, & O’Garra, 2020; Sarvaria, Madrigal, & Saudemont, 2017).

The role of pro-tumor TANs in association with disease prognosis was further verified using TCGA database. The patients with higher expression of gene panels specific to pro-tumor TANs exhibited a poorer prognosis compared to those with lower expression (Fig. 5F). These findings suggest that the increased malignancy observed in ADC, particularly in HPV-negative cases, may also be attributed to the enrichment of pro-tumor TANs and evasion of anti-tumor TANs.

Cellular heterogeneity of plasma/B cells in ADC

In our study, we identified 6 distinctive clusters of plasma/B cells based on the set of gene enrichment: Plasma/B_01(marked with IGHA2, IGHG2 and IGLC2), Plasma/B_02 (marked with HLA-DRA, HLA-DPB and CD37), Plasma/B_03 (marked with KRT17 and S100A2), Plasma/B_04 (marked with CCL5, CXCL13, IL-32 and CCL4), Plasma/B_05 (marked with CXCL8), Plasma/B_06 (marked with HMGB2) (Fig. 6A-C). Among these clusters, the most abundant one was Plasma/B_01_IGHA2 (Fig. 6D and Supplementary Fig. 6A-B). Interestingly, the proportion of Plasma/B_01_IGHA2 was found to be significantly higher in ADC compared to SCC, and it was also slightly higher in HPV-negative cases than in HPV-positive cases (Fig. 6D and Supplementary Fig. 6A). Generally, B-lymphocytes are known to have tumor-inhibitory properties. However, the ectopic gene module of plasma/B cells in ADC suggested that some sub-clusters might present a tumor-promoting role. This hypothesis was further supported by the results of GOBP analysis, which indicated that Plasma/B_01_IGHA2 gene signatures were more active in negatively regulating B cell activation, complement activation and receptor signaling pathway (Fig. 6E). When we combined the top 50 signature genes in Plasma/B_01_IGHA2-enriched subset and performed the survival analysis using TCGA database, we observed that higher expression of the signature genes predicted a poorer prognosis of CC patients (Fig. 6F). Taken together, the cellular heterogeneity of plasma/B cells in the TIME of ADC is well identified at the single-cell level.

Phenotype diversity of plasma/B cells in ADC.

A. UMAP plots showing the distribution of plasma/B cells in 15 samples (left) and the re-clustering into 6 clusters (right) according to ectopic expressions of genes. B. Plot heatmap showing the annotation of each sub-cluster with top 5 DEGs. C. UMAP plots demonstrating the most specific genes as the marker for each sub-cluster. D. UMAP and histogram plots to compare the differences of distribution and proportion of each plasma/B cell sub-cluster between different histological types. E. GOBP analysis of immunity-related signaling pathways that are more active in ADC, in terms of Plasma/B_01_IGHA2 sub-cluster. The histogram data are transformed from -Log10 (P-value). F. Kaplan-Meier curve showing the overall survival rate of CC patients stratified by the top 50 genes-scaled signature of Plasma/B_01_IGHA2.

Crosstalk among tumor cells, Tregs and neutrophils establishes the immunosuppressive TIME in ADC

In the above description, we have outlined that certain ADC-enriched cell sub-clusters, such as Epi_10_CYSTM1, Tregs, and pro-tumor TANs, played oncogenic roles in regulating various cellular activities, including maintenance of stemness and evasion of immune surveillance. To gain deeper insights into the interplay among these clusters and their influence on ADC progression, we harnessed the Cellchat tool (Jin et al., 2021). Our primary focus revolved around unraveling the mechanisms governing Treg recruitment to the tumor microenvironment and the maintenance of stemness within ADC-enriched cell sub-clusters.

Given that the 3 ADC-specific sub-clusters (Epi_07_CAPS, Epi_10_CYSTM1 and Epi_12_RRAD) of epithelial cells exhibited quite high levels of stem-like properties, and particularly that Epi_10_CYSTM1 presented the highest degree of malignancy among them, we further analyzed their interactions with other types of cells. Firstly, in comparison to SCC, ADC-enriched epithelial sub-clusters exhibited a higher propensity for interaction with Tregs through ligand-receptor signaling pathways, including ALCAM (ALCAM>>CD6) and MHC-II (HLA-DR>>CD4) (Fig. 7A-E), which have been reported to be essential for Treg recruitment, expansion and stabilization to establish the immunosuppressive microenvironment (Chalmers et al., 2022; Ferragut, Vachetta, Troncoso, Rabinovich, & Elola, 2021; Freitas et al., 2019). To further uncover the underlying reason for the enhanced stemness observed in ADC epithelial cells, we focused on the signaling received by Epi_10_CYSTM1 cell cluster. Interestingly, the Cellchat analysis revealed that, compared to other pathways, such as CD46>>JAG1 and GZMA>>PARD3, the interaction of TGFβ>>TGFBR between Tregs and epithelial cells (particularly in cluster Epi_10_CYSTM1) was remarkedly activated in ADC but not in SCC (Fig. 7F-G). TGFβ is a canonical secreted protein that induces carcinogenesis, such as cancer cell proliferation, invasion and self-renewal (Derynck, Turley, & Akhurst, 2021; Massagué, 2008; Tauriello, Sancho, & Batlle, 2022). Thus, we speculate that the recruited Tregs might secrete TGFβ to stimulate the tumor epithelial cells of ADC, leading to increased stemness. Additionally, pro-tumor TANs were also recruited to the tumor area and interacted with Epi_10_CYSTM1, as well as Epi_07_CAPS and Epi_12_RRAD cells, via the ANNEXIN-to-FPR1/FPR2 signaling pathway (Supplementary Fig. 7A-B). Previous studies have reported that the ANNEXIN pathway can promote the chemotaxis of neutrophils (Araújo et al., 2021). Likewise, TANs expressing isg, which have been shown to promote tumorigenesis as mentioned earlier, were also recruited to the tumor area through the ANNEXIN pathway. These findings suggest that the aggressive clusters of TANs may contribute to the formation of an ADC-specific TIME.

The cellular interaction modules of sub-clusters of T cells, neutrophils and tumor epithelial cells.

A. Circle plots showing the interacting networks between epithelial cell sub-clusters and T cell sub-clusters via the pathway of ALCAM, by comparing ADC and SCC. B. Circle plots simplified from A to show the interactions among target cell clusters via ALCAM pathway. C. Circle plots showing the interacting networks between epithelial cell sub-clusters and T cell sub-clusters via the pathway of MHC-II, by comparing ADC and SCC. D. Circle plots simplified from C to show the interactions among target cell clusters via MHC-II pathway. From A to D, The direction of each arrow shows the regulation from outputting cells to incoming cells. The width of the each line shows the predicted weight and strength of regulation. E. Bubble plot showing the probability of ligand-to-receptor combination of each pathway between two different target sub-clusters of cells, by comparing ADC with SCC. F. Circle plots showing Tregs regulate epithelial cells via the TGF-β pathway, which is solely activated in ADC. G. Bubble plot showing the probability of ligand-to-receptor combination of TGF-β pathway between Tregs and epithelial cells, by comparing ADC with SCC. The pathways of ADGRE5, CD46, GZMA and NAMPT are used as negative controls. H. Dual IF staining confirming that in SLC26A3high regions of CC tissues, more FOXP3+ cells are recruited than in SLC26A3low regions (left). The numbers of recruited FOXP3+ cell are quantified using histogram plot (right). Three individual samples with ROI were calculated and P<0.01 were marked with **, showing significant difference. I. Multiplexed IF staining confirming the interaction between CD6 (on FOXP3+ cells) and ALCAM (on SLC26A3+ epithelial cells) in the ALCAM pathway. J. Multiplexed IF staining showing that the recruitment of FOXP3+ cells towards SLC26A3+ cells might induce EMT (marked with E-cadherin) and increase the stemness (marked with ALDH1A1) of tumor cells, via TGF-β pathway.

To approve these hypotheses, we conducted immunofluorescence (IF) analysis and observed a greater recruitment of FOXP3+ cells in close proximity to regions exhibiting high SLC26A3 expression (Epi_10_CYSTM1 cell cluster) (Fig. 7H). Furthermore, when compared to regions with low SLC26A3 expression, tumors displaying high SLC26A3 levels exhibited significantly increased expression of Aldehyde Dehydrogenase 1 Family Member A1 (ALDH1A1), a known marker for cancer stem cells in CC (Douville, Beaulieu, & Balicki, 2009) (Supplementary Fig. 7C). Notably, we identified SLC26A3high tumor cells expressing ALCAM in close proximity to CD6+ Tregs, suggesting potential involvement of ALCAM>>CD6 signaling in Treg recruitment (Fig. 7I). Additionally, we observed reduced E-cadherin expression in tumor cells within SLC26A3high regions, possibly induced by TGFβ secreted by adjacent Tregs (Fig. 7J and Supplementary Fig. 7D). Altogether, these findings confirm the intricate crosstalk between the Epi_10_CYSTM1 cell cluster and Tregs, which plays a pivotal role in establishing an immunosuppressive TIME and sustaining heightened stemness in tumor cells.

Discussion

Cervical cancer is often regarded as less aggressive compared to ovarian cancer and endometrial cancer, primarily due to the effectiveness of HPV vaccines in prevention. When diagnosed at early stages, the 5-year survival rate for cervical cancer patients exceeds 90%, surpassing that of other gynecologic malignancies (Gennigens et al., 2022; Schiffman, Castle, Jeronimo, Rodriguez, & Wacholder, 2007). However, if cervical cancer metastasizes to the intraperitoneal or retroperitoneal lymph nodes, the survival rate will decrease to 30%∼60% (Cohen et al., 2019). Worse still, the 5-year survival rate for recurrent CC patients is less than 20% (H. Li, Wu, & Cheng, 2016). Specifically, when considering the ADC type alone, the prognosis is even worse. Unfortunately, our understanding for ADC is limited. Most previous studies have concentrated on SCC type of CC, and obtained heterogenetic information derived from bulk RNA-seq method. Therefore, it is worthwhile to shift our attention towards an in-depth investigation on the disparities between ADC and SCC at the single-cell resolution.

More recently, researchers have started to utilize scRNA-seq approach to investigate the tumor microenvironment of CC. For instance, Cao et al compared 5 cases of SCC tumor tissues with paired adjacent normal tissues and described the immune landscape in which specific clusters of T and B cells exhibited immune-exhausting or activating processes (Cao et al., 2023). Another study conducted by Qiu et al aimed to elucidate the distinct molecular patterns of immune reactions between ADC and SCC in the context of different HPV infection status at the single-cell level (Qiu et al., 2023). However, due to limited samples tested in these studies, our knowledge on the TIME in CC is yet inadequate. In the current study, we have generated a genomic atlas of ADC, which may depict the landscape of TIME built by both ADC tumor cells and neighboring cells. Apart from the predominant presence of epithelial cells, the cell clusters in ADC are majorly composed of T cells, tumor-associated neutrophils (TANs), and plasma/B cells. The gene subsets identified within ADC-abundant cell sub-clusters, including Epi_10_CYSTM1, Tregs, pro-tumor TANs and P/B_01_IGHA2, exert potential roles in promoting both the carcinogenesis and immunosuppression of ADC. In the TIME of ADC, the unique module of cellular interaction has been decoded and validated that the immunosuppressive Tregs and the tumor-promoting TANs are recruited to tumor cells via chemo-attraction. Finally, our hypothesis from scRNA-seq data is further confirmed using clinical samples, providing evidence that SLC26A3 is a reliable biomarker for ADC (Supplementary Fig. 8).

It has been reported that immunotherapies are less effective on ADC patients, especially those who are not infected with HPV (Attademo et al., 2020; Ferrall, Lin, Roden, Hung, & Wu, 2021). As the two major components of ADC cell types, T cells and epithelial cells have been shown to be the dominant regulators in the TIME due to their reciprocal interaction dynamics. We identified several ADC-specific epithelial cell clusters that exhibit strong stemness and aggressiveness. Among these clusters, Epi_10_CYSTM1 demonstrated the highest level of malignancy, but the underlying mechanism behind this behavior remains unknown. The Cellchat analysis revealed that Tregs might quite actively crosstalk to Epi_10_CYSTM1. To confirm this, immunostaining for co-localization demonstrated an increased aggregation of Tregs around tumor cells exhibiting overexpression of SLC26A3. It has also been widely reported that tumor cells acquire higher stemness present more aggressive features of resistance to immunotherapy (Dianat-Moghadam et al., 2022). This study confirmed that TGFβ produced by Tregs induce the stemness of Epi_10_CYSTM1 cells by activating downstream oncogenic pathways, inducing EMT on tumor epithelial cells (Nixon, Gao, Wang, & Li, 2023). Interestingly, Epi_10_CYSTM1 cells express ALCAM, which is a soluble ligand, and bind to its receptor CD6 on Tregs, activating chemotaxis and facilitating Treg recruitment. The infiltration of a large number of Tregs in ADC significantly accelerates immune evasion (Supplementary Fig. 8). Therefore, the presence of a higher number of Tregs infiltrating the tumor area of ADC may explain the increased resistance to immunotherapy. It is worth noting that, in comparison to the proportions of Tregs, the majority of T cell fractions in the CC tumor area are comprised of cytotoxic T cells and exhausted T cells. However, their distribution is relatively equal, indicating the dysfunctional status of the majority of infiltrated T cells in the CC tumor area.

Another interesting phenomenon observed in our sing-cell sequencing data is the involvement of TANs in the regulatory interactions within the TIME of ADC. It has been clearly investigated that pro-tumor TANs, characterized by CD66, CD11, CD170 and PD-L1, exert the oncogenic roles by promoting tumor cell proliferation, angiogenesis, inducing genetic instability, and most importantly causing immunosuppression (Jaillon et al., 2020). In this study, pro-tumor TANs are predominantly enriched in ADC tissues compared to SCC, especially in HPV-negative ADC cases. This suggests that ADC-specific epithelial cell sub-clusters actively engage in the recruitment of pro-tumor TANs through the ANXA1-FPR1/FPR2 ligand-receptor interaction.

The anti-tumor function of B cells has been identified previously (Bod et al., 2023; Helmink et al., 2020). However, our study has presented a specific cluster of plasma/B cells with tumor-promoting functions. The overexpression of gene signatures from Plasma/B_01_IGH2 cells is associated with poor regression in CC patients. Interestingly, Plasma/B_01-IGHA2 was found to be closely related to the PC Cluster (Plasma cell Cluster) in Cao’s study due to the shared gene set (Cao et al., 2023), such as IGHA1, IGHA2, IGHG2, IGHG4, etc. However, their roles were contradictory. According to that study, PC cluster was derived from the local expansion of TIME and showed potential effectiveness in anti-tumor immunity. In our case, Plasma/B_01-IGHA2 had a reverse effect and played a tumor-promoting role in ADC. This difference in function is likely attributed to the distinct cell types and their interactions in ADC, instead of SCC. Although the cell-to-cell interaction did not yield positive results, probably owing to the small cell proportion, the underneath dynamics of B cells still need further investigation.

Most importantly, in this study SLC26A3 is identified as a promising diagnostic biomarker to predict lymph node metastasis in ADC patients at stage IIIC. In 2018, the International Federation of Gynecology and Obstetrics (FIGO) updated a new criterion of stage IIIC by including lymph node status, which indicates that the therapeutic strategy is completely different from early stages. There are two stratification of stage IIIC: IIICp (p stands for pathological findings of LN, which means misdiagnosis) and IIICr (r stands for radiological findings of LN, which means pre-surgical evidence) (Bhatla, Aoki, Sharma, & Sankaranarayanan, 2018). Unfortunately, radiological imaging tools do have some limitations, as CT or MRI can only detect suspiciously metastatic LN when the minimum diameter is ≥ 10mm (Saleh et al., 2020). However, over 10% of CC patients who are initially diagnosed with early-stage and deemed suitable for radical surgery, end up being upstaged due to lymph node metastasis discovered after the surgery (Dabi et al., 2018; Thelissen et al., 2022). For these patients, they require additional radiotherapy such as external beam irradiation besides surgery (Abu-Rustum et al., 2020). However, if the presence of lymph node metastasis could have been accurately diagnosed in biopsy sample, the optimal therapeutic approach for these patients would be radiotherapy alone. Moreover, combination of both surgery and radiotherapy may lead to more complications for patients than radiotherapy alone, without offering additional benefits (Cushman et al., 2018; Kashima et al., 2023; J. Yang et al., 2015). To solve this problem, a reliable biomarker is required to diagnose LN metastasis in conjunction with radiographic tool. Combining two large cohorts of clinical data, we identified that CC patients with ADC types were more vulnerable to be misdiagnosed for clinical staging. Derived from scRNA-seq data, SLC26A3 could be employed as a marker for the Epi_10_CYSTM1 cluster, which was found to be exclusively enriched in stage IIIC cases of ADC. The IHC staining further confirmed that ADC patients at stage IIIC exhibited higher expression of SLC26A3 compared to early stages (FIGO stage I∼ II). Furthermore, we tested on biopsy samples and it implied that SLC26A3 could be a significant predictor of LN metastasis. Our study for the first time provides evidence that ADC patients are at a higher risk of developing lymph node metastasis, which can be challenging to detect solely through radiological imaging techniques. However, by incorporating lymph node imaging with biomarkers such as SLC26A3, it may significantly reduce the misdiagnosis rate among FIGO IIICp patients.

In summary, we characterized the genomic landscape of the TIME in ADC at a single-cell resolution. Our study revealed the presence of specific epithelial cell clusters in ADC that exhibited high aggressiveness. The enriched gene signatures within these clusters may serve as potential therapeutic targets to improve the efficacy of immunotherapy. The interactions between tumor epithelial cells and other types of cells, such as T cells and neutrophils, comprehensively depict a highly immunosuppressive microenvironment of ADC. Moreover, our investigation extended to the issue of upstaging at the single-cell level, with the identification of potential biomarkers from specific tumor cell clusters showing promise as diagnostic indicators for the precise diagnosis of cervical cancer patients.

Additional information

eLife Assessment

This manuscript is useful to researchers with an interest in cervical cancers because it provides scRNA-seq data from a diverse cohort of 15 early-stage cervical cancer patients. While the dataset could be of use to the research community, the key claims of the paper around the immunosuppressive microenvironment associated with specific tumour cell clusters (and the properties/importance of those clusters) are incomplete. Additional experiments will be required to substantiate these claims.

Reviewer #1 (Public Review):

Summary:

The authors in this manuscript performed scRNA-seq on a cohort of 15 early-stage cervical cancer patients with a mixture of adeno- and squamous cell carcinoma, HPV status, and several samples that were upstaged at the time of surgery. From their analyses they identified differential cell populations in both immune and tumour subsets related to stage, HPV status, and whether a sample was adenocarcinoma or squamous cell. Putative microenvironmental signaling was explored as a potential explanation for their differential cell populations. Through these analyses the authors also identified SLC26A3 as a potential biomarker for later stage/lymph node metastasis which was verified by IHC and IF. The dataset is likely useful for the community, however, the strong claims made are not adequately supported by the data and would require additional functional validation.

Strengths:

The dataset could be useful for the community.

SLC26A3 could potentially be a useful marker to predict lymph node metastasis with further study.

Weaknesses:

The link between the background in the introduction and the actual study and findings is often tenuous or not clearly explained. A re-working of the intro to better set up and link to the study questions would be beneficial.

For the sequencing, which kit was used on the Novaseq6000?

Additional details are needed for the analysis pipeline. How were batch effects identified/dealt with, what were the precise functions and settings for each step of the analysis, how was clustering performed and how were clusters validated etc. Currently, all that is given is software and sometimes function names which are entirely inadequate to be able to assess the validity of the analysis pipeline. This could alternatively be answered by providing annotated copies of the scripts used for analysis as a supplement.

For Cell type annotation, please provide the complete list of "selected gene markers" that were used for annotation.

No statistics are given for the claims on cell proportion differences throughout the paper (for cell types early, epithelial sub-clusters later, and immune cell subsets further on). This should be a multivariate analysis to account for ADC/SCC, HPV+/- and Early/Late stage.

The Y-axis label is missing from the proportion histograms in Figure 2D. In these same panels, the bars change widths on the right side. If these are exclusively in ADC, show it with a 0 bar for SCC, not doubling the width which visually makes them appear more important by taking up more area on the plot.

Throughout the manuscript, informatic predictions (differentiation potential, malignancy score, stemness, and trajectory) are presented as though they're concrete facts rather than the predictions they are. Strong conclusions are drawn on the basis of these predictions which do not have adequate data to support. These conclusions which touch on essentially all of the major claims made in the manuscript would need functional data to validate, or the claims need to be very substantially softened as they lack concrete support. Indeed, the fact that most of the genes examined that were characteristic of a given cluster did not show the expected expression patterns in IHC highlights the fact that such predictions require validation to be able to draw proper inferences.

The cluster Epi_10_CYSTM1 which is the basis for much of the paper is present in a single individual (with a single cell coming from another person), and heavily unconnected from the rest of the epithelial populations. If so much emphasis is placed on it, the existence of this cluster as a true subset of cells requires validation.

Claims based on survival analysis of TCGA for Epi_10_CYSTM1 are based on a non-significant p-value, though there is a slight trend in that direction.

The claim "The identification of Epi_10_CYSTM1 as the only cell cluster found in patients with stage IIICp raises the possibility that this cluster may be a potential marker to diagnose patients with lymph node metastasis." This is incorrect according to the sample distributions which clearly show cells from the patient who has EPI_10_CYSTM1 in multiple other clusters. This is then used as justification for SLC26A3 which appears to be associated with associated with late stage, however, in the images SLC26A3 appears to be broadly expressed in later tumours rather than restricted to a minor subset as it should be if it were actually related to the EPI_10_CYSTM1 cluster.

The authors claim that cytotoxic T cells express KRT17, and KRT19. This likely represents a mis-clustering of epithelial cells.

Multiple claims are made for specific activities based on GO term biological process analysis which while not contradictory to the data, certainly are by no means the only explanation for it, nor directly supported.

Reviewer #2 (Public Review):

Summary:

Peng et al. present a study using scRNA-seq to examine phenotypic properties of cervical cancer, contrasting features of both adenocarcinomas (ADC) and squamous cell carcinoma (SCC), and HPV-positive and negative tumours. They propose several key findings: unique malignant phenotypes in ADC with elevated stemness and aggressive features, interactions of these populations with immune cells to promote an immunosuppressive TME, and SLC26A3 as a biomarker for metastatic (>=Stage III ) tumours.

Strengths:

This study provides a valuable resource of scRNA-seq data from a well-curated collection of patient samples. The analysis provides a high-level view of the cellular composition of cervical cancers. The authors introduce some mechanistic explanations of immunosuppression and the involvement of regulatory T cells that are intriguing.

Weaknesses:

I believe that many of the proposed conclusions are over-interpretations or unwarranted generalizations of the single-cell analysis. These conclusions are often based on populations in the scRNA-seq data that are described as enriched or specific to a given group of samples (eg. ADC). This conclusion is based on the percentage of cells in that population belonging to the given group; for example, a cluster of cells that dominantly come from ADC. The data includes multiple samples for each group, but statistical approaches are never used to demonstrate the reproducibility of these claims.

This leads to problematic conclusions. For example, the "ADC-specific" Epi_10_CYSTM1 cluster, which is a central focus of the paper, only contains cells from one of the 11 ADC samples and represents only a small fraction of the malignant cells from that sample (Sample 7, Figure 2A). Yet, this population is used to derive SLC26A3 as a potential biomarker. SLC26A3 transcripts were only detected in this small population of cells (none of the other ADC samples), which makes me question the specificity of the IHC staining on the validation cohort.

This is compounded by technical aspects of the analysis that hinder interpretation. For example, it is clear that the clustering does not perfectly segregate cell types. In Figures 2B and D, it is evident that C4 and C5 contain mixtures of cell type (eg. half of C4 is EPCAM+/CD3-, the other half EPCAM-/CD3+). These contaminations are carried forward into subclustering and are not addressed. Rather, it is claimed that there is a T cell population that is CD3- and EPCAM+, which does not seem likely.

Author response:

The following is the authors’ response to the original reviews.

Public reviews:

Reviewer #1 (public review):

(1) The link between the background in the introduction and the actual study and findings is often tenuous or not clearly explained. A re-working of the intro to better set up and link to the study questions would be beneficial.

We have rewritten the introduction of the manuscript and clearly stated the study questions we were aiming for:

In paragraph 1-we have stated clearly that we need to study why ADC type of cervical cancer is more aggressive. (Line 58 - 77)

In paragraph 2- we have stated clearly that we need to find valuable biomarkers to help diagnose lymph node metastasis, which may compensate the shortage of radiological imaging tools and reduce the rate of misdiagnosis. (Line 78 - 100)

In paragraph 3- we have stated clearly that HPV negative cases is a special group of cervical cancer and we aim to study its cellular features. (Line 101 - 108)

In paragraph 4- we have stated clearly that we need to decode cell-to-cell interaction mode in the tumor immune microenvironment of ADC using scRNA-seq. (Line 109 - 123)

(2) For the sequencing, which kit was used on the Novaseq6000?

For sequencing, we used the Chromium Controller and Chromium Single Cell 3’Reagent Kits (v3 chemistry CG000183) on the Novaseq6000. We feel sorry for lacking this quite important part and have already add the information in Methods section. (Line 196- 197)

(3) Additional details are needed for the analysis pipeline. How were batch effects identified/dealt with, what were the precise functions and settings for each step of the analysis, how was clustering performed and how were clusters validated etc. Currently, all that is given is software and sometimes function names which are entirely inadequate to be able to assess the validity of the analysis pipeline. This could alternatively be answered by providing annotated copies of the scripts used for analysis as a supplement.

We apologize for the inadequacy of descriptions of data analysis process. We have already provided a new part of “data processing” with more details in the Methods section (Line 202 - 221). In addition, we have also provided annotated copies of scripts in the supplementary data as Supplementary Data 1.

(4) For Cell type annotation, please provide the complete list of "selected gene markers" that were used for annotation.

We have already added the list of marker genes for cell type annotation in the revised manuscript as Supplementary Table 3.

(5) No statistics are given for the claims on cell proportion differences throughout the paper (for cell types early, epithelial sub-clusters later, and immune cell subsets further on). This should be a multivariate analysis to account for ADC/SCC, HPV+/- and Early/Late stage.

We feel sorry for lacking statistics when performing analyses of comparisons. In the revision, we have already used statistic approaches to analyze the differences between each set of group comparison. As a result, the corresponding figures have been revised, accordingly.

For examle, Fig. 1F, Fig. 2D, Fig. 4E, Fig. 5D, Fig. 6D had been re-analyzed to compare ADC/SCC；Supplementary Fig. 1A, Supplementary Fig. 2A, Supplementary Fig. 4A, Supplementary Fig. 5A, Supplementary Fig. 6A had been re-analyzed to compare HPV+/HPV-; Supplementary Fig. 1B, Supplementary Fig. 2B, Supplementary Fig. 4B, Supplementary Fig. 5B, Supplementary Fig. 6B had been re-analyzed to compare Early/Late stage. All P values have been listed in the figure legends.

(6) The Y-axis label is missing from the proportion histograms in Figure 2D. In these same panels, the bars change widths on the right side. If these are exclusively in ADC, show it with a 0 bar for SCC, not doubling the width which visually makes them appear more important by taking up more area on the plot.

We feel sorry for impreciseness when presenting histograms of Fig. 2D and we have also revised other figures with similar mistakes, such as Fig. 1F, Fig. 5D. As for the width of bars, which is due to output style of data processing, we have already corrected all similar mistakes alongside the whole manuscript, for example, Fig. 2D and Supplementary Fig. 2A-B.

(7) Throughout the manuscript, informatic predictions (differentiation potential, malignancy score, stemness, and trajectory) are presented as though they're concrete facts rather than the predictions they are. Strong conclusions are drawn on the basis of these predictions which do not have adequate data to support. These conclusions which touch on essentially all of the major claims made in the manuscript would need functional data to validate, or the claims need to be very substantially softened as they lack concrete support. Indeed, the fact that most of the genes examined that were characteristic of a given cluster did not show the expected expression patterns in IHC highlights the fact that such predictions require validation to be able to draw proper inferences.

Thank you for your insightful comments. As you noted, several conclusions were initially based on bioinformatics predictions. Thus in the revised manuscript, we have rewritten all relevant descriptions in a more softened way, particularly in the paragraph of “epithelial cells” in Results section, as well as the conclusions derived from bioinformatics predictions in other paragraphs throughout the manuscript. We hope our revised descriptions will enhance the precision of our work.

For example, in paragraph “The sub-clusters of epithelial cells in ADC exhibit elevated stem-like features (from Line 353)”, many over-affirmative disriptions had been re-written in Line 353, 362, 371, 375, 379, 383, 390, 392. From Line 395 to 399, the conclusion had been revised as “The observation of cluster Epi_10_CYSTM1 and its possible specificity to ADC makes us question whether or not it may be related to the aggressiveness of ADC” compared to the previous “This observation may partially indicate that high stemness cluster Epi_10_CYSTM1 is essential for ADC to present more aggressive features”. From Line 400 to 408, conclusions from GO analyses had also been rewritten.

In paragraph “ADC-specific epithelial cluster-derived gene SLC26A3 is a potential prognostic marker for lymph node metastasis (from Line 422)”, many conclusions based on predictions had been revises, such as Line 424 - 428, Line 439 - 441, Line 451 - 453, Line 455 - 457, Line 458 - 459, Line 471 - 473, Line 478 - 481, Line 484 - 486, Line 489, etc.

In paragraph “Tumor associated neutrophils (TANs) surrounding ADC tumor area may contribute to the formation of a malignant microenvironment (from Line 536)”, we have changed the descriptions based on bio-infomative predictions, such as Line 560, Line 561, Line 565, Line 566, Line 572, Line 576 - 577, etc.

In paragraph “Crosstalk among tumor cells, Tregs and neutrophils establishes the immunosuppressive TIME in ADC (from Line 601)”, we have already corrected the all the affirmative descriptions, such as Line 604, Line 612, Line 614, Line 626, Line 628 - 629, Line 641, Line 654 – 655, etc.

All the changes have also been listed in Revision Notes in detail.

(8) The cluster Epi_10_CYSTM1 which is the basis for much of the paper is present in a single individual (with a single cell coming from another person), and heavily unconnected from the rest of the epithelial populations. If so much emphasis is placed on it, the existence of this cluster as a true subset of cells requires validation.

We appreciate this suggestion. We agree that the majority of Epi_10_CYSTM1 cells are derived from sample S7. The fact that we have detected this cluster in only one patient may be due to sampling differences and the inherent heterogeneity of tumor specimens. However, the relatively high number of cells in this cluster from one stage III patient suggests its presence in ADC patients and highlights its potential as a diagnostic marker for clinical staging. To further investigate whether this cluster is generally existing in ADC patients, we have identified and selected candidate genes, such as SLC26A3, ORM1, and ORM2, as representative markers of this cluster, which demonstrated high specificity (as shown in Fig. 3B). We then performed IHC staining on a total of 56 tissue samples, and the results showed positive expressions of these markers in the majority of stage IIIC tumor tissues, confirming the existence of this cell cluster (as shown in Supplementary Fig. 3E). In our revised manuscript, we have included an in-depth discussion of this issue in the seventh paragraph of the Discussion section (From Line 801).

(9) Claims based on survival analysis of TCGA for Epi_10_CYSTM1 are based on a non-significant p-value, though there is a slight trend in that direction.

Thank you for your insightful comment. From the data of TCGA survival analysis for Epi_10, we found a not-so-slight trend of difference between groups (with a small P value). As a result, we presented this data and hoped to add more strength to the clinical significance of this cluster. However, this indeed caused controversy because the P value is non-significant. As a result, we have already deleted this data in the revised manuscript.

(10) The claim "The identification of Epi_10_CYSTM1 as the only cell cluster found in patients with stage IIICp raises the possibility that this cluster may be a potential marker to diagnose patients with lymph node metastasis." This is incorrect according to the sample distributions which clearly show cells from the patient who has EPI_10_CYSTM1 in multiple other clusters. This is then used as justification for SLC26A3 which appears to be associated with associated with late stage, however, in the images SLC26A3 appears to be broadly expressed in later tumours rather than restricted to a minor subset as it should be if it were actually related to the EPI_10_CYSTM1 cluster.

We feel thankful for this question. The conclusion that “The identification of Epi_10_CYSTM1 as the only cell cluster found in patients with stage IIICp raises the possibility that this cluster may be a potential marker to diagnose patients with lymph node metastasis” has indeed been written too concrete according to the sample distribution. We feel sorry for this and have already corrected the description into “As one of stage IIIC-specific cell clusters, the cluster of Epi_10_CYSTM1, with its representative marker gene SLC26A3, presents potential diagnostic value to predict lymph node metastasis” from Line 478-481.

However, based on our results, we do think this cluster is a potential diagnostic marker and the hypothesis is right. As for SLC26A3, we have specifically added a new paragraph (from Line 801 - 822) in Discussion section to discuss the rationality and necessity of selecting this gene as our central focus, and the reasons why SLC26A3 should be the representative of cluster Epi_10_CYSTM1. As you noted, SLC26A3 appears to be broadly expressed in later tumors rather than restricted to a minor subset in the images. We apologize for any misunderstanding caused. When presenting the IHC data, we only showed the strongly positive areas of each slide to emphasize the differences. In our revision, we have included whole slide scanning images of the IHC samples, clearly showing that SLC26A3 is restricted to a part of the tumors (Supplementary Fig.9).

(11) The authors claim that cytotoxic T cells express KRT17, and KRT19. This likely represents a mis-clustering of epithelial cells.

We apologize for using data without noticing the contamination of T cells with few epithelial cells. We have re-performed quality control to exclude contamination and re-analyzed all data of T cells. In the reviesed manuscript, we have therefore updated completely new data for T cells in both Fig. 4 and Supplementary Fig. 4.

(12) Multiple claims are made for specific activities based on GO term biological process analysis which while not contradictory to the data, certainly are by no means the only explanation for it, nor directly supported.

Our initial purpose was to use GO analysis as supports for our conclusions. However, we know these are only claims but not evidence, which is also the problem of our writing techniques as in question (7). Therefore, in our revised manuscript, we have already deleted GO data and descriptions in the paragraphs of “T cell (Fig.4)”(from Line 495) and “B/plasma cell (Fig.6)” (from Line 579), because the predictions are quite irrelevant to our conclusions.

However, in the sections of “epithelial cell (Fig.2)” (from Line 352) and “neutrophils (Fig.5)” (from Line 536), we retained the GO data and rewrote the conclusions, because these analyses have provided us with valuable information regarding the role of specific cell clusters in ADC progression. Furthermore, our subsequent analyses, such as CellChat, have further validated the accuracy of the findings from the GO analysis. We do think this logically supports the whole storyline of the study.

Reviewer #2 (public review):

(1) I believe that many of the proposed conclusions are over-interpretations or unwarranted generalizations of the single-cell analysis. These conclusions are often based on populations in the scRNA-seq data that are described as enriched or specific to a given group of samples (eg. ADC). This conclusion is based on the percentage of cells in that population belonging to the given group; for example, a cluster of cells that dominantly come from ADC. The data includes multiple samples for each group, but statistical approaches are never used to demonstrate the reproducibility of these claims.

We feel sorry that many of the conclusions have been written in an over-affirmative way but lack profound supporting evidences. In our revision, we have already optimized the writing techniques and re-written all conclusions or descriptions related to only bio-informatic predictions. Moreover, we have performed statistical re-analyses on all data and rearranged the related figures.

For example, in Line 352, we have changed the sub-title “The sub-clusters of epithelial cells exhibit elevated stem-like features to promote the aggressiveness of ADC” into “The sub-clusters of epithelial cells in ADC exhibit elevated stem-like features”. In this paragraph, many over-affirmative discriptions such as “exclusively”, “significant”, “overwhelmingly”, “remarkably” have been deleted. From Line 486-493, the conclusion of “Moreover, SLC26A3 could be employed as a marker for the Epi_10_CYSTM1 cluster, aiding in the diagnosis of lymph node metastasis to prevent post-surgical upstaging in ADC patients in the future” have been changed into “our results propose that SLC26A3 might be considered as a diagnostic marker to predict lymph node metastasis in ADC patients”. Similar over-affirmative descriptions and conclusions had also been re-written in the other paragraphs, which has been refered to question (7) above.

(2) This leads to problematic conclusions. For example, the "ADC-specific" Epi_10_CYSTM1 cluster, which is a central focus of the paper, only contains cells from one of the 11 ADC samples and represents only a small fraction of the malignant cells from that sample (Sample 7, Figure 2A). Yet, this population is used to derive SLC26A3 as a potential biomarker. SLC26A3 transcripts were only detected in this small population of cells (none of the other ADC samples), which makes me question the specificity of the IHC staining on the validation cohort.

We sincerely feel grateful for this question. This is a quite important question as it is also pointed out by reviewer#1 in question (8) above. In the revised manuscript, we have already optimized our descriptions and have added detailed explanation for the importance of SLC26A3 in the Discussion section (from Line 802 - 823). We agree that the majority of Epi_10_CYSTM1 cells are derived from sample S7. The fact that we detected this cluster in only one patient may be due to sampling differences and the inherent heterogeneity of tumor specimens. However, the relatively high number of cells in this cluster from one stage III patient suggests its presence in ADC and highlights its potential as a diagnostic marker for staging ADC. To further investigate whether this cluster is generally present in ADC patients, we identified and selected candidate genes, such as SLC26A3, ORM1, and ORM2, as representative markers of this cluster, which demonstrated high specificity (as shown in Fig. 3B). We then performed IHC staining on 56 cases of tissue samples, and the results showed positive expression of these markers in the majority of stage III tumor tissues, confirming the existence of this cell cluster (as shown in Supplementary Fig. 3E). In our revised manuscript, we have included an in-depth discussion of this issue in the seventh paragraph of the Discussion section.

(3) This is compounded by technical aspects of the analysis that hinder interpretation. For example, it is clear that the clustering does not perfectly segregate cell types. In Figures 2B and D, it is evident that C4 and C5 contain mixtures of cell type (eg. half of C4 is EPCAM+/CD3-, the other half EPCAM-/CD3+). These contaminations are carried forward into subclustering and are not addressed. Rather, it is claimed that there is a T cell population that is CD3- and EPCAM+, which does not seem likely.

Thank you for your insightful comment. This important point is also raised by reviewer#1 above. In the revised manuscript, we have reanalyzed our scRNA-seq data and listed the canonical marker genes for cell type annotation. Most importantly, as for T cells and its sub-clustering, we have performed quality control and re-analyzed all data for T cells, with contamination excluded. In the reviesed manuscript, we have added the re-analyzed data for T cells in both Fig. 4 and Supplementary Fig. 4.

Recommendations for the authors:

Reviewer #1 (recommendations for the authors):

The text would substantially benefit from an editorial revision of language usage.

We sincerely feel grateful for this suggestion. In our revision, we have conducted language editing and carefully rewritten our manuscript. The changes have been clearly marked in the tracked version of the revised manuscript.

Reviewer #2 (recommendations for the authors):

(1) Use statistical approaches to claim enrichment/specificity of populations to given groups (ADC, HPV, etc). Analysis packages like Milo for differential abundance testing would be very helpful.

We feel grateful for this suggestion. In our revision, we have performed statistical analyses for all groups of comparison data. Meanwhile, we have rearranged the figures based on these statistical results, for example, Fig. 1F, Fig. 2D, Fig. 4E, Fig. 5D, Fig. 6D, Supplementary Fig. 1A-B, Supplementary Fig. 2A-B, Supplementary Fig. 4A-B, Supplementary Fig. 5A-B, Supplementary Fig. 6A-B.

(2) In the subclustering, consider a round of quality control to ensure that all cells are of the cell type they are claimed to be. Contaminant clusters/cells could be filtered out or reassigned. This could be supplemented with an automated annotation approach using cell-type references.

We feel thankful for this suggestion. As a result, we have provided copies of scripts in the supplementary data to ensure the quality control of cell type annotation.

(3) An explanation for why SLC26A3 is so rare in the scRNA-seq data, but seemingly common in the IHC staining would be helpful. I am concerned about the specificity of the stain.

We apologize for lacking adequate explanation of SLC26A3 and cluster Epi_10_CYSTM1. This is a quite crucial question as it has been listed above in question (8) of reviewer #1 and question (2) of reviewer #2 (public review section). In the revised manuscript, we have added intenstive discussion about this question in the seventh paragraph of Disccusion section (from Line 801 - 822). In fact, because of the heterogeneity among different individuals and different tumor regions even within one sample, Epi_10_CYSTM1 seemed to be derived from only one sample. However, the relatively high number of cells in this cluster from one late-stage (stage IIIC) patient suggests its presence in ADC and highlights its potential as a diagnostic marker for staging ADC. Furthermore, we have identified SLC26A3, ORM1 and ORM2 as specific markers of this cluser and performed IHC staining. With a positive expression of these markers, the existence of this cluster has been indirectly proved (as shown in Fig. 3B).

Version 2

Abstract

Abstract

Background

Cervical adenocarcinoma (ADC) is more aggressive compared to other types of cervical cancer (CC), such as squamous cell carcinoma (SCC). The tumor immune microenvironment (TIME) and tumor heterogeneity are recognized as pivotal factors in cancer progression and therapy. However, the disparities in TIME and heterogeneity between ADC and SCC are poorly understood.

Methods

We performed single-cell RNA sequencing on 11 samples of ADC tumor tissues, with other 4 SCC samples served as controls. The immunochemistry and multiplexed immunofluorescence were conducted to validate our findings.

Results

Compared to SCC, ADC exhibited unique enrichments in several sub-clusters of epithelial cells with elevated stemness and hyper-malignant features, including the Epi_10_CYSTM1 cluster. ADC displayed a highly immunosuppressive environment characterized by the enrichment of regulatory T cells (Tregs) and tumor-promoting neutrophils. The Epi_10_CYSTM1 cluster recruits Tregs via ALCAM-CD6 signaling, while Tregs reciprocally induce stemness in the Epi_10_CYSTM1 cluster through TGFβ signaling. Importantly, our study revealed that the Epi_10_CYSTM1 cluster could serve as a valuable predictor of lymph node metastasis for CC patients.

Conclusions

This study highlights the significance of ADC-specific cell clusters in establishing a highly immunosuppressive microenvironment, ultimately contributing to the heightened aggressiveness and poorer prognosis of ADC compared to SCC.

Introduction

Cervical cancer (CC) is the fourth leading cause of cancer-related death among all types of women malignancies (Sung et al., 2021). The two major histological types of CC are squamous cell carcinoma of cervix (SCC) and adenocarcinoma of cervix (ADC), accounting for 70% and 25% of all cases, respectively (Cohen, Jhingran, Oaknin, & Denny, 2019). Compared to SCC, ADC patients have a higher rate of rapid progression, relapse and insensitivity to chemotherapy, radiotherapy and even immunotherapy. Therefore, ADC is considered as a more aggressive phenotype than SCC (Sasieni, Castanon, & Cuzick, 2009). Unfortunately, the current standard treatment for CC is not stratified based on pathologic types. As a result, ADC patients often have suffered from therapeutic failures and experienced short-term recurrence and metastasis, even finishing standard treatment procedures (Sasieni et al., 2009). However, due to the low morbidity of ADC, in-depth investigations are insufficient. Furthermore, the molecular mechanisms elucidating its aggressiveness and the associated biomarkers are not yet well understood. Therefore, this study aims to comprehensively compare the intratumor heterogeneity of ADC, hoping to find out the potential molecule markers specifically associated with ADC.

For early-stage CC patients, it is a crucial and unavoidable issue to have post-surgical upstaging. Over 10% of patients, who are initially diagnosed with early stage and deemed suitable for radical surgery, end up being upstaged due to lymph node metastasis discovered after the surgery (Dabi et al., 2018; Thelissen et al., 2022). This indicates misdiagnosis, leading to different therapeutic strategies and outcome. According to NCCN (National Comprehensive Cancer Network) guidelines, these misdiagnosed patients require additional radiotherapy after surgery, even though they have underwent surgery (Bhatla, Aoki, Sharma, & Sankaranarayanan, 2018). However, combo-treatment of surgery and radiotherapy may cause additional complications and heavier financial burdens, without offering more benefits than radiotherapy alone (Cushman et al., 2018; Kashima et al., 2023). Actually, the optimal therapeutic approach should be radiotherapy alone if the presence of lymph node metastasis could have been accurately diagnosed beforehand (Abu-Rustum et al., 2020). To avoid this problem, the International Federation of Gynecology and Obstetrics (FIGO) in 2018 updated a new stage as IIIC by including lymph node (LN) status and stratified this stage into two sub-groups: IIICp (p stands for pathological findings of LN, which means misdiagnosis) and IIICr (r stands for radiological findings of LN, which means pre-surgical evidence) (Bhatla et al., 2018). Unfortunately, radiological imaging tools do have limitations, since CT or MRI can only detect suspiciously metastatic LN when the minimum diameter is ≥ 10mm (Saleh et al., 2020). Therefore, together with radiological diagnosis, it is urgent to search for potential molecular predictors of LN metastasis, in order to decrease the rate of post-surgical upstaging.

Persistent infection with human papillomavirus (HPV) is the primary cause of CC (Mann, Singh, & Kumar, 2023). However, there are about 5%-15% of patients testing negative for HPV (Yang et al., 2023). It is reported that HPV-negative patients tend to experience worse clinical outcomes (Fernandes, Viveros-Carreño, Hoegl, Ávila, & Pareja, 2022). However, current therapeutic strategies for CC do not take into account the HPV infection status as a factor for stratified therapy. Therefore, it is imperative to investigate the precise mechanisms underlying the development of HPV non-infected tumors.

The tumor immune microenvironment (TIME), consisting of tumor cells, immune cells, cytokines and their interactions, predominantly determines the efficacy of immunotherapy (Lv et al., 2022). In cervical cancer, while it has been reported that 30%-50% of patients exhibit positive PD-L1 expression, however, only an estimated 20%-30% are expected to respond positively to PD-1/PD-L1 blockade therapy (Omenai, Ajani, & Okolo, 2022; Sun et al., 2020). Therefore, cervical cancer, especially the pathological type of ADC, is characterized as an immune-insensitive cancer type, due to atypical modulation of the TIME (Cao et al., 2023). The mechanisms of tumor immune evasion is basically ascribed to dysregulation of TIME and are similar across different types of cancer, involving the deactivation of immune surveillance pathways or recruitment of immunosuppressive cells (Vinay et al., 2015). However, the specific characteristics of the TIME in ADC remain poorly understood. Therefore, in order to optimize precise therapeutic strategies and to enhance the efficacy of immunotherapy, it is indeed imperious to investigate the intricacies of TIME in ADC.

To extensively investigate the cellular and molecular heterogeneity of CC, particularly ADC, the utilization of single-cell RNA sequencing (scRNA-seq) technique is crucially needed. In the current study, we focused on the remodeling of TIME in ADC and performed scRNA-seq analysis on 11 ADC samples, with another 4 SCC samples included as controls. By high-throughput analysis, we are able to characterize the landscape of TIME in ADC, with different types of tumor-promoting cells being recruited in the vicinity of the tumor, thus creating an immunosuppressive environment in ADC. The cellular crosstalk analysis provides valuable insights into the interactions between different cell types within the TIME. Furthermore, we have conducted a screening of novel signature genes as prognostic biomarkers for ADC. These biomarkers have the potential to possibly predict lymph node metastasis and may guide personalized treatment decisions more effectively.

Methods

Clinical sample collection

The 15 cases of fresh cervical cancer tissues were collected from Hunan Cancer Hospital, with informed consent obtained from 15 independent patients. The inclusion criteria are as follows: 1) The diagnosis of CC, with pathological type of ADC or SCC, should be confirmed by pathologists; 2) Pre-surgical FIGO stages should be within the range from IB (which means the tumor mass should be detectable via CT or MRI) to IIA (which means no parametrial involvement, nor distant metastasis), in which patients have explicit indication for radical surgery following NCCN guidelines; 3) No neo-adjuvant treatments should be given before the surgery, including chemotherapy and radiotherapy; 4) The surgery options should be type-C radical hysterectomy and include pelvic lymphadenectomy with/without para-aortic lymphadenectomy. Detailed information of the 15 patients was listed in Supplementary Table 1. This study was supervised and approved by the Ethics Committee Board of Hunan Cancer Hospital.

Sample preparation for single-cell isolation

Immediately after surgical removal, the fresh tumor tissues were washed by saline for 3 times and stored in cold storage solution (MACS® Tissue Storage Solution) for transportation. To prepare high-quality samples, the tissues were cut into pieces to 1mm3 and were incubated at 37°C for 30 minutes. 0.25% trypsin solution was used to digest the tissue pieces at 37°C for 10 minutes. Then single-cell samples were filtered through a 70-μm meshed filter and suspended in a red blood cell lysis buffer (MACS® Red Blood Cell Lysis Solution) for dissolution. Before instrumental sequencing, suspension of single cells were visualizing by TC20TM Automated Cell Counter to assess the cell viability. Each sample was confirmed to have a cell viability rate of over 80%.

Single-cell RNA sequencing

The suspension of single cells was transferred onto the 10X Chromium Single-Cell instrument to generate single-cell beads in the emulsion (GEMs). Then the scRNA-seq libraries were constructed by using the Chromium Controller and Chromium Single Cell 3’Reagent Kits (v3 chemistry CG000183) and sequenced by using the sequencer Novaseq6000 (Illumina, USA). All procedures were following the manufacturer’s protocol.

Data processing, dimension reduction and cell clustering

The raw scRNA-seq reads were processed for barcode processing, genome mapping and the gene expression matrixes were generated by using the Cell Ranger toolkit (version 5.0.0) of 10X Genomics platform. The GRCh38 human reference genome was utilized in the read alignment process. The unique molecular identifiers (UMIs) were counted in each cell. As for data processing, cells with low qualities, which showed <200 expressed genes or >25% mitochondrial UMIs, were excluded. The package of Seurat R (version 4.0.5 R) was applied for quality control. The Scrublet software (version 0.2.2) was utilized to identify and remove potential doublets. As for data normalization and conversion, the LogNormalize method was implemented in the NormalizeData function, and then the normalized counts were log-transformed. With the function of RunPCA, the principal component analysis (PCA) was used for data dimension reduction. The functions of FindNeighbors and FindClusters were used for cell clustering. For visualization, the RunUMAP function was performed. In each unsupervised cell cluster, the gene markers were identified by the function of FindAllMarkers with comparison to other clusters.

Cell type annotation

The cell typing was conducted and annotated according to the selected gene markers via CellMarker database and publications (C. Li et al., 2022; Qiu et al., 2023; Xue et al., 2022). The expression of differential expressed genes (DEGs) was used to determine each cell type following the rules: Log2Foldchange should be > 0.2 and adjusted-P values should be < 0.05.

Pseudotime trajectory analysis

To identify the developmental changes of epithelial cells, the pseudotime trajectory analyses were performed by using Monocle2 R package (version 2.22.0), in which Seurat was used as input. Calculated by the Monocle algorithm, the top 50-1000 genes were conducted for differential GeneTest. Then, the DDR-Tree and default parameters were generated to visualize the translational relationship and developmental orders among different epithelial sub-clusters.

Cell malignancy scoring

Among these epithelial cells, the malignancy score of each single cell was predicted by using scCancer R package (Guo et al., 2021). The malignant cells were distinguished from non-malignant cells by inferring large-scale copy-number variations (CNVs), which was calculated by the interCNV R package as described (Guo et al., 2021; Xue et al., 2022).

Signaling pathway enrichment analysis

The biological signaling pathway of each cell cluster was identified by performing Gene Ontology (GO) analyses and Hallmark Pathway enrichment analyses, according to the Molecular Signature Database (MSigDB version 7.5.1). Then DEGs of each cell cluster were converted to ClusterProfiler (version 4.2.2) package for analyses of predicted functions.

CytoTRACE

In order to predict the differential state of cells, scRNA-seq derived data were pass to the CytoTRACE software (version 0.3.3) with R package for further analyses.

Survival analysis

The CESC dataset from TCGA database was downloaded from Xena, with clinical characteristics, survival endpoints and treatments were included. The Kaplan-Meier survival curves were generated by using R software.

Cellular communication analysis

The possible interactions, with ligand-to-receptor communicating signaling pathway, between two types of cells were analyzed using the R packages of CellChat (version 1.5.0). Seurat was used for normalization and the rank with significance was calculated. The probability of ligand-to-receptor interaction in each signaling pathway was plotted by P-value and intensity.

Immunohistochemistry (IHC)

The paraffin-embedded ADC samples (whether post-surgical or biopsy samples) were sectioned into 4μm slides. The protocol of IHC strictly followed our previous published works (Peng et al., 2019). The primary antibody of SLC26A3 was purchased from Santa Cruz Biotechnology (Cat# sc-376187) and diluted at the ratio of 1:20. The protein expression of SLC26A3 was determined using the method of H-scoring: H-score = intensity score × percentage of positivity. The intensity was scored as: Negative=0, Weak=1, Moderate=2, Strong=3. Samples were grouped as the higher expression group (H-score ≥ 150) and the lower expression (H-score < 150) group of SLC26A3.

Multiplexed immunofluorescence (multi-IF)

The multi-IF was performed using the Opal 6-Plex Manual Detection Kit (NEL811001KT, Akoya), following the manufacturer’s instructions. Generally, the slides were stained with each primary antibody sequentially following the staining steps in each cycle as: (1) antigen retrieval by citrate (pH=6) at water bath for 30 minutes; (2) independent primary antibody incubation at 4°C atmosphere overnight or 37°C atmosphere for 1 hour; (3) common HRP-crosslinked secondary antibody incubation at 37°C atmosphere for 30 minutes; (4) independent opal reactive fluorophore solution at intended wavelength (opal 570 was excluded in order to save for FOXP3 signal). Each independent primary antibody started a new cycle. After all cycles finished, the slides were prepared for incubation of FOXP3 monoclonal antibody with eFluor™ 570 (Thermo Fisher, Cat#41-4777-82) conjugated, as well as DAPI (Sigma) for nucleus staining.

Statistics

Specific analyses of scRNA-seq data were decoded and visualized by using the R software. Statistics were processed using IBM-SPSS Statistics 20.0 software and R software as well, including Student’s t-test, two-sided Wilcoxon test, etc. Correlations analysis was performed by the Chi-square test. Survival analysis was performed by the Kaplan-Meier method. P values < 0.05 were considered as statistically significant.

Results

Single-cell transcriptomic landscape of cervical cancer

We collected 15 samples of cervical cancer tissues from 15 individual patients immediately after undergoing radical hysterectomy surgery. Among these cases, 11 were ADC types, with 5 being HPV-negative and 6 being HPV-positive. The remaining 4 cases were SCC types, with an even distribution of HPV-positive and HPV-negative status (Supplementary Table 1). Prior to surgery, all 15 patients were diagnosed as early FIGO stages (stage I ∼IIA) based on evaluation of CT, MRI, and physical examination. However, after surgery, 5 cases turned out to be upstaged with pathological evidence of lymph node metastasis (i.e. FIGO stage IIICP), requiring theses patients to receive radiation therapy according to NCCN guidelines. Therefore, in addition to investigate the difference between histological types and HPV infection status, we also focused on the issue of post-surgical upstaging by comparing different FIGO stages.

In order to analyze the genomic profiling, we resolved these samples at the single-cell level using the 10X genomics chromium platform. The cell viability and sample quality were ensured beforehand, and a total of 102,012 qualified cells were utilized for further analysis (Fig. 1A). The technique of Uniform Manifold Approximation and Projection (UMAP) was utilized to reduce the dimension of gene profiling data for visualization. The obtained cells were primarily categorized into 12 cell clusters, which were further re-clustered into 9 distinct cell types, based on specific gene markers (C. Li et al., 2022) (Fig. 1B-C). These cell types included T cells (38,627 cells in total, proportioning 37.87%, marked with PTPRC, CD3D, CD3E and CD3G), epithelial cells (32,199 cells in total, proportioning 31.56%, marked with EPCAM, SLP1 and CD24), neutrophils (12,586 cells in total, proportioning 12.34%, marked with CSF3R), macrophages (6,798 cells in total, proportioning 12.34%, marked with CD68 and CD163), fibroblasts (4,258 cells in total, proportioning 4.17%, marked with COL1A2 and DCN), plasma cells (4,044 cells in total, proportioning 3.96%, marked with JCHAIN), endothelial cells (1,389 cells in total, proportioning 1.36%, marked with ENG and VWF), B cells (1,326 cells in total, proportioning 1.30%, marked with CD79A and MS4A1) and mast cells (785 cells in total, proportioning 0.77%, marked with MS4A2) (Fig. 1D-E and Supplementary Table 2-3).

The single-cell genomic atlas of cervical cancer.

A. The schematic design of sample collection, single-cell RNA sequencing, data processing and clinical validating through 10X genomics platform (The sketch of scRNA-sq machine, as well as the plot of dimension-reduction visualization, was originally cited from the official website of 10X genonics: https://www.10xgenomics.com). B. UMAP plots of all single cells by original clustering and cell type re-clustering, according to the published lineage-specific marker genes, respectively. C. Violin plots demonstrating the expression of marker genes that correspond to each of the major cell types. D. UMAP plots demonstrating the most typical marker gene specific to each type of cell cluster. E. UMAP plot and histogram plot showing the occupation ratios of the major cell types in each individual sample, with HPV status and histological type summarized. F. UMAP (up) and histogram (down) plots to show the differences of distribution and proportion of each cell type between different histological types (ADC vs. SCC). Statistics were performed using R software with two-sided Wilcoxon test (P values for each group are listed below: T cell: P=0.661; Epithelial cell: P=0.039; Neutrophil: P=0.026; Macrophage: P=0.661; Fibroblast: P=0.661; Plasma: P=0.078; Endothelial cell: P=0.851; B cell: P=0.104; Mast cell: P=0.753). Statistics were shown as *P < 0.05; **P < 0.01; n.s. not significant.

According to our scRNA-seq data, ADC exhibited a significant higher proportion of neutrophils than SCC (20.05% vs. 4.23%). The proportion of T cells was also higher in ADC (41.13%) compared to SCC (35.48%). Conversely, SCC showed a nearly threefold enrichment of epithelial cell (47.43%) compared to ADC (16.48%) (Fig. 1F). When considering the HPV infection status, HPV-negative patients demonstrated decreased proportions of epithelial cells (7.78% vs. 28.38%) and neutrophils (16.66% vs. 24.68%) compared to HPV-positive patients. The proportion of T cells enriched in HPV-negative cases (45.77%) was higher than that in HPV-positive cases (32.42%). HPV-negative patients exhibited prominently higher proportions of plasma cells (10.13% vs. 1.99%) and B cells (2.84% vs. 0.77%) compared to HPV-positive patients, even though their proportions were lower than T cells, epithelial cells and neutrophils (Supplementary Fig. 1A). Furthermore, we conducted a comparison between early-stage and late-stage cases, revealing that late-stage patients have a higher proportion of epithelial cells (35.50%) than early-stage cases (28.82%) (Supplementary Fig. 1B). In summary, these data describes primarily a complex microenvironment in cervical cancer, especially in the ADC type, with varied composition of immune and tumor cells under different conditions.

The sub-clusters of epithelial cells in ADC exhibit elevated stem-like features

The aforehand data has shown that the majority of the CC tumor is composed of epithelial cells. In order to explore the characteristics of these epithelial cells within the TIME of ADC, we further classified them into 12 sub-clusters based on the subsets of differently expressed genes (DEGs) in each group (Fig. 2A). Each sub-cluster was named based on the most prominently up-regulated gene in each panel of signature genes (Fig. 2B-C). Several epithelial cell clusters were strongly enriched in ADC than SCC, such as Epi_04_TFF2, Epi_06_TMC5, Epi_07_CAPS and Epi_08_SCGB3A1. Notably, Epi_09_SST, Epi_10_CYSTM1, Epi_11_REG1A and Epi_12_RRAD seemed to be exclusively enriched in ADC, not in SCC (Fig. 2D). The stratified enrichment of different cluster between ADC and SCC has captured our attention for further investigation. Additionally, Epi_09_SST, Epi_10_CYSTM1 and Epi_11_REG1A were solely enriched in HPV-positive cases, although no distinctive clusters had been found in HPV-negative patients (Supplementary Fig. 2A). When comparing samples from patients in the early stages with those who have lymph node metastasis, 3 clusters (Epi_02_IGLC2, Epi_05_CCL5 and Epi_12_RRAD) were slightly increased among the late-stage patients. Surprisingly, we found that the sub-cluster Epi_10_CYSTM1 was exclusively present in stage IIIC patients, indicating that it might be a potential target for identification of the biomarkers for late stage patients (Supplementary Fig. 2B).

The scRNA-seq data reveals the malignant features of tumor epithelial cells.

A. UMAP plots showing the distribution of epithelial cells in 15 samples and the sub-clustering into 12 clusters according to ectopic gene expressions. Each cluster of epithelial cells was named using the most highly enriched gene. B. Heatmap plot showing the annotation of each epithelial sub-cluster with top 5 DEGs. C. UMAP plots demonstrating the most specific gene as the marker for each sub-cluster. D. UMAP (left) and histogram (right) plots to compare the differences of distribution and proportion of each epithelial cell sub-cluster between different histological types (ADC vs. SCC). Statistics were performed using R software with two-sided Wilcoxon test (P values for each group are listed below: Epi_01_NTS: P=0.040; Epi_02_IGLC2: P=0.661; Epi_03_TMRPSS11E: P=0.661; Epi_04_TTF2: P=0.104; Epi_05_CCL5: P=0.950; Epi_06_TMC5: P=0.412; Epi_07_CAPS: P=0.040; Epi_08_SCGB3A1: P=0.412; Epi_09_SST: P<0.001; Epi_10_CYSTM1: P<0.001; Epi_11_REG1A: P<0.001; Epi_12_RRAD: P<0.001). Statistics were shown as *P < 0.05; **P < 0.01; n.s. not significant. E. UMAP plot of CytoTRACE showing the thermal imaging projection of predicted developmental order (left) and histogram plot showing the ranking of CytoTRACE scores (right). F. UMAP plot of malignancy analysis demonstrating the cell malignancy features (up) and ranking of scores (down). G. GOBP analyses showing the top 15 enriched signaling pathways of Epi_10_CYSTM1 that are more active in ADC than SCC (up), particularly pathways related to cell-to-cell interactions (down). The histogram data are transformed from -Log10(P-value) for visualization.

Abbreviations: Ig: immunoglubolin; MHC: major histocompatibility complex.

We further evaluated the features of each epithelial cluster, especially their predicted roles in regulating tumorigenesis. To assess the differentiation potential of each cell cluster, we employed the Cellular Trajectory Reconstruction Analysis using gene Counts and Expression (CytoTRACE) method (Fig. 2E). The results showed that clusters of Epi_07_CAPS, Epi_09_SST, Epi_10_CYSTM1 and Epi_12_RRAD ranked the top four in CytoTRACE score, indicating the potential for pluripotency and differentiation, also referred to stem-like features. Interestingly, Epi_09_SST, Epi_10_CYSTM1 and Epi_12_RRAD were exclusively enriched in ADC. The proportion of cluster Epi_07_CAPS, which acquired the highest level of stemness, was higher in ADC, although Epi_07_CAPS was also identified in SCC (Fig. 2D). Subsequently, pseudotime analysis was conducted to elucidate the differentiation trajectories. The results showed that Epi_07_CAPS, Epi_09_SST and Epi_10_CYSTM1 were positioned towards the end of pseudotime developmental trajectory, which was in consistence with their high stem-like characteristics and indicated a higher degree of malignancy in ADC than SCC (Supplementary Fig. 2C). In addition, we performed the malignancy scoring analysis, which demonstrated that cluster Epi_10_CYSTM1 was predicted to exhibit the highest degree of malignancy compared to other clusters (Fig. 2F). Interestingly, cluster Epi_10_CYSTM1 displayed a diverse developmental profile and was exclusively identified in ADC. Conversely, Epi_07_CAPS and Epi_12_RRAD, which exhibited high stem-like characteristics, were ranking lower in terms of malignancy scoring when compared to other clusters. The observation of cluster Epi_10_CYSTM1 and its potential specificity to ADC raises the question of whether this cluster is associated with the aggressiveness of ADC.

Besides, we performed Gene Ontology Biological Process (GOBP) analysis on specific signaling pathways based on the gene expression pattern of cluster Epi_10_CYSTM1 (Fig. 2G). The results revealed that cluster Epi_10_CYSTM1 tended to be more active in regulating immunity-related pathways, which indicated this cluster might be possibly associated with dysfunction of immunity. In addition, cluster Epi_10_CYSTM1 was predicted to be more active in regulating cell-to-cell interaction (Fig. 2G). Therefore, further investigation is warranted to explore the intricate crosstalk networks between epithelial cells and other cell types in the TIME of ADC.

ADC-specific epithelial cluster-derived gene SLC26A3 is a potential prognostic marker for lymph node metastasis

Based on the aggressive characteristics of cluster Epi_10_CYSTM1 from above bioinformatic predictions, we further examined the correlation between this cluster and the clinical features of CC. For validation, we alternatively examined cluster Epi_12_RRAD (Supplementary Fig. 3A), which was another malignant cluster predominantly found in ADC (Figure. 2D). We firstly checked the DEGs of these two clusters. In Epi_12_RRAD cluster, we identified Insulin-like growth factor 2 (IGF2) and Alcohol dehydrogenase 1C (ADH1C), which were exclusively expressed in Epi_12_RRAD, as two most specific genes of this cluster (Supplementary Fig. 3B). However, the immunohistochemistry (IHC) staining showed that IGF2 was almost negative across cases with different clinical stages (Supplementary Fig. 3C). On the other hand, ADH1C was strongly expressed in both early-stage and late-stage cases (Supplementary Fig. 3C). Likewise, the survival analysis of Epi_12_RRAD-enriched gene signatures did not demonstrate a significant difference between the two groups (Supplementary Fig. 3D). These results suggest that cluster Epi_12_RRAD may not be a potential target when relating to the clinical features of ADC.

In the cluster of Epi_10_CYSTM1, the DEG with the highest expression is CYSTM1 (Fig. 3A). However, CYSTM1 was also expressed in other cell clusters and showed low specificity (Fig. 3B). On the other hand, SLC26A3, ORM1 and ORM2 were specifically enriched in Epi_10_CYSTM1 and could be considered as potential candidate biomarkers for representation of this cluster. Nevertheless, ORM1/OMR2 were positively expressed but were not satisfied to distinguish the severity of CC cases, due to an irrelevance between expression intensity and clinical stages (Supplementary Fig. 3E). Interestingly, SLC26A3, as a representative marker of cluster Epi_10_CYSTM1, showed potential value to associated with late clinical stages for CC patients (Supplementary Fig. 3E). This is encouraging, since currently it lacks biomarkers to predict late stages of cervical cancer and the only way to detect LN metastasis beforehand is using radiological tools, which have technical limitations.

SLC26A3 is identified as a potential prognostic and diagnostic indicator for lymph node metastasis of CC patients.

A. Heatmap showing the top 50 DEGs of Epi_10_CYSTM1 cluster in comparison to all the other sub-clusters (Red: Epi_10_CYSTM1 cluster; Green: other epithelial cell sub-clusters). B. Violin plots showing the expression difference between two groups (top), with UMAP plots (bottom) demonstrating the specificity of each candidate gene, filtering SLC26A3 as the most identical marker for this sub-cluster. C. IHC staining showing the protein expression of SLC26A3 in surgically-resected ADC samples which are classified as early stages (FIGO stage I-IIA) and late stages (FIGO stage IIIC1∼2p). Images from 6 individual cases were shown as representatives for each group. D. IHC staining showing the protein expression of SLC26A3 in biopsy ADC samples which are classified as early stages (FIGO stage I-IIA) and late stages (FIGO stage IIIC1∼2p). Images from 4 individual cases were shown as representatives for each group. The method of H-score is used and the scoring system is as follows: negative (0), weak (1), intermediate (2), and strong (3). Expression is quantified by the H-score method.

This is a problem because misdiagnosis of clinical staging affects the disease outcome and treatment approach for CC patients. In order to examine the fact of post-surgical upstaging, we merged data from two independent patient cohorts obtained from two clinical centers: Xiangya hospital (Cohort 1) and Hunan cancer hospital (Cohort 2). The criterion for objective selection was outlined in Supplementary Fig. 3F. The rates of upstaging for each center are 10.85% (23 out of 212 in Cohort 1) and 15.31% (214 out of 1398 in Cohort 2), respectively. These rates fall within the range reported by previous studies (Dabi et al., 2018; Thelissen et al., 2022). Interestingly, both cohorts of data revealed that the misdiagnose rate was significantly higher in HPV-negative patients than HPV-positive patients, especially among ADC patients. From the data of Cohort 2, it was observed that ADC patients were more likely to be misdiagnosed, which means that metastatic LNs were more difficult to detect in ADC patients via radiological tools. In Cohort 1, we have also observed the same tendency although without significant difference between different histological types, probably due to a smaller sample size (Table. 1-2).

As one of stage IIIC-specific (Supplementary Fig. 2B) cell clusters, Epi_10_CYSTM1, with its representative marker gene SLC26A3, presents potential diagnostic value to predict lymph node metastasis, which has been shown above (Fig. 3B and Supplementary Fig. 3E). Therefore, we try to validate SLC26A3’s association with staging of IIIC, via detecting the expression pattern on post-surgical ADC samples, and more practically on biopsy samples. The results from IHC staining showed that ADC patients with stage IIIC had a higher expression of SLC26A3 (Fig. 3C, D and Table 3, 4). In summary, our results propose that SLC26A3 might be considered as a diagnostic marker to predict lymph node metastasis in ADC patients.

Enrichment of regulatory T cells indicates an immunosuppressive status of ADC

T cells play a crucial role as the primary effectors of tumor immunity. Our data reveals that T cells constitute the predominant cell population within the TIME of cervical cancer (Fig. 1F). A total of 52,297 T cells were categorized into 8 main clusters (Fig. 4A). The gene expression patterns of T cells in CC were in accord with those observed in other types of cancers (Sorin et al., 2023; Xue et al., 2022). Based on canonical genes (C. Li et al., 2022), these cells were re-clustered as the following 4 types of T cells: exhausted T cells (marked with CD8, TIGIT, PDCD1, LAG3, HAVCR2), cytotoxic T cells (marked with CD8 and CD3), regulatory T cells (Treg, marked with CD4, FOXP3 and IL2RA), activated T cells (marked with CD8, C69 and CD3G) (Fig. 4A-C). The subsets of DEGs from each type of T cells have been shown in Supplementary Fig. 4C. When comparing the checkpoint pathway states among different T cell clusters, we observed relatively higher levels of LAG3 and PDCD1 (PD1) in exhausted T cells, while CTLA4 and TIG1 were more highly expressed in Tregs. These genes were identified to present immunosuppressive functions (Qiu et al., 2023) (Fig. 4D). On the other hand, genes such as ICOS and TNFRSF, which indicate the activation of immune checkpoint pathways, exhibited low expression levels in cytotoxic T cells and activated T cells (Fig. 4D). These findings lead us to hypothesize that the establishment of an immunosuppressive TIME in ADC may involve the recruitment of regulatory T cells (Tregs) to the tumor area and the subsequent inactivation of a substantial proportion of cytotoxic T cells.

Cellular and molecular heterogeneity of T cells in ADC.

A. UMAP plots showing the sample distribution (left), original cell clustering of T cells (middle, 8 clusters in total) and re-clustering of T cell subtypes (right, 4 major sub-clusters: exhausted T cell, cytotoxic T cell, regulatory T cell and activated T cell) according to acknowledged marker genes. B. Violin plot showing the expression of specific marker genes that annotate each sub-type of T cells. C. UMAP plots showing the widely recognized classification markers that denote each type of T cell sub-cluster. D. Dot heatmap plots that demonstrate the level of marker genes representing the signaling pathways of immune checkpoint activation and inhibition for grouped sub-clusters. E. Differences of distribution and proportion of each T cell sub-cluster between different histological types (ADC vs. SCC). Statistics were performed using R software with two-sided Wilcoxon test (P values for each group are listed below: Exhausted T: P=0.571; Cytotoxic T: P=0.078; Treg: P<0.01; Activated T: P=0.040). Statistics were shown as *P < 0.05; **P < 0.01; n.s. not significant.

Abbreviation: Treg: regulatory T cell.

We further compared different status of CC with different attribution of T cell subtypes. It was demonstrated that ADC had a significantly higher proportion of Tregs compared to SCC (Fig. 4E). Meanwhile, ADC patients who tested negative for HPV showed a higher enrichment of Tregs and a lower proportion of activated T cells compared to those with HPV-positive status (Supplementary Fig. 4A). These findings shed light on the underlying mechanism for immunotherapy insensitivity of ADC, especially in HPV-negative status.

Tumor associated neutrophils (TANs) surrounding ADC tumor area may contribute to the formation of a malignant microenvironment

The large-scale enrichment of specific gene subsets in neutrophils in CC prompted us to further examine their roles. A total of 12,586 neutrophils were detected and the composition proportion in ADC (20.05%) was almost five times higher than that in SCC (4.23%) (Fig. 1F). Based on this observation, we hypothesize that ADC-specific neutrophil clusters may be related to poor prognosis and increased malignancy. All the cells were initially grouped into 6 clusters based on the gene expression modules (Fig. 5A). In the context of cancer, tumor associated neutrophils (TANs) can exhibit dual functions, either promoting tumor (pro-tumor TANs) or inhibiting tumor progression (anti-tumor TANs). Based on the consensus markers identified by Jaillon et al. (Jaillon et al., 2020) and Xue et al.(Xue et al., 2022), we performed further clustering of the neutrophils in our dataset, resulting in the identification of 3 sub-types of TANs: pro-tumor TANs (Neu_03, marked with CD66, CD11 and MT1X), anti-tumor TANs (Neu_02, marked with CD66, CD11, CCL5, KRT5 and ELF3), TANs with isg (interferon stimulated genes) and other undefined types (Neu_01, distinguished by CD66, CD11, IFITM2, S100A8 and LST1) (Fig. 5B-C). Generally, the Neu_01 sub-cluster, which consists of TANs with isg with the subset of specific genes, accounted for the largest proportion among all neutrophils. Notably, the presence of pro-tumor TANs was significantly higher in ADC compared to SCC (Fig. 5D), as well as in HPV-negative cases compared to HPV-positive cases (Supplementary Fig. 5A). On the other hand, the proportion of anti-tumor TANs in ADC was predominantly lower than that in SCC (Fig. 5D).

The heterogeneity of tumor associated neutrophils in ADC.

A. UMAP plots showing the distribution of neutrophils in 15 samples (left), original clustering (middle) the re-clustering (right) into 3 sub-clusters according to gene markers of tumor-associated neutrophils (TANs). B. Dot heatmap showing the top 5 DEGs in each sub-cluster of TANs. C. UMAP plots for annotation of each sub-type of TANs with published marker genes presented: pro-tumor TANs, anti-tumor TANs and TANs with isg. D. UMAP and histogram plots to compare the differences of distribution and proportion of each sub-types of TANs between different histological types (ADC vs. SCC). Statistics were performed using R software with two-sided Wilcoxon test (P values for each group are listed below: TANs with isg: P=0.002; anti-TANs: P=0.002; pro-TANs: P=0.049). Statistics were shown as *P < 0.05; **P < 0.01; n.s. not significant. E. GOBP analyses of signaling pathways that are more active in ADC than SCC, in terms of the pro-tumor TANs cluster. The histogram data are transformed from -Log10 (P-value). F. Kaplan-Meier curve showing the overall survival rate of CC patients stratified by the top 90 genes-scaled signature of pro-TANs.

We further conducted the GOBP analysis on gene subsets derived from pro-tumor TANs to predict their potential functions. The findings indicated that pro-tumor TANs in ADC might be more active in regulating cell-to-cell interactions and chemotaxis. Specifically, they were involved in pathways such as chemokine-mediated or cytokine-mediated signaling, neutrophil activation, and receptor internalization (Fig. 5E). These results suggest that pro-tumor TANs may engage in communication with various cell types. Interestingly, pro-tumor TANs were predicted to acquire enhanced activation in mediating inhibitory pathways related to immunity (Fig. 5E). These pathways include the negative regulation of leukocyte proliferation and lymphocyte activation, as well as the suppression of IL-10 production, which is known to promote immune responses (Saraiva, Vieira, & O’Garra, 2020; Sarvaria, Madrigal, & Saudemont, 2017).

The role of pro-tumor TANs in association with disease prognosis was further investigated using TCGA database. The patients with higher expression of gene panels specific to pro-tumor TANs exhibited a poorer prognosis compared to those with lower expression (Fig. 5F). These findings suggest that the increased malignancy observed in ADC, particularly in HPV-negative cases, may also be related to the enrichment of pro-tumor TANs.

Cellular heterogeneity of plasma/B cells in ADC

In our study, we identified 6 distinctive clusters of plasma/B cells based on the set of gene enrichment: Plasma/B_01(marked with IGHA2, IGHG2 and IGLC2), Plasma/B_02 (marked with HLA-DRA, HLA-DPB and CD37), Plasma/B_03 (marked with KRT17 and S100A2), Plasma/B_04 (marked with CCL5, CXCL13, IL-32 and CCL4), Plasma/B_05 (marked with CXCL8), Plasma/B_06 (marked with HMGB2) (Fig. 6A-C). Among these clusters, the most abundant one was Plasma/B_01_IGHA2 (Fig. 6D and Supplementary Fig. 6A-B). Interestingly, the proportion of Plasma/B_01_IGHA2 was found to be higher in ADC compared to SCC, and it was also slightly higher in HPV-negative cases than in HPV-positive cases (Fig. 6D and Supplementary Fig. 6A). Generally, B-lymphocytes are known to have tumor-inhibitory properties. However, the ectopic gene module of plasma/B cells in ADC suggested that some sub-clusters might present a tumor-promoting role. When we combined the top 50 signature genes in Plasma/B_01_IGHA2-enriched subset and performed the survival analysis using TCGA database, we observed that higher expression of the signature genes predicted a poorer prognosis of CC patients (Fig. 6E). Taken together, the cellular heterogeneity of plasma/B cells in the TIME of ADC is identified at the single-cell level.

Phenotype diversity of plasma/B cells in ADC.

A. UMAP plots showing the distribution of plasma/B cells in 15 samples (left) and the re-clustering into 6 clusters (right) according to ectopic expressions of genes. B. Plot heatmap showing the annotation of each sub-cluster with top 5 DEGs. C. UMAP plots demonstrating the most specific genes as the marker for each sub-cluster. D. UMAP and histogram plots to compare the differences of distribution and proportion of each plasma/B cell sub-cluster between different histological types. Statistics were performed using R software with two-sided Wilcoxon test (P values for each group are listed below: Plasma/B_01_IGHA2: P=0.040; Plasma/B_02_HLA-DAR: P=0.177; Plasma/B_03_KRT17: P=0.040; Plasma/B_04_CCL5: P=0.388; Plasma/B_05_CXCL8: P=0.169; Plasma/B_06_HMGB2: P=0.211). Statistics were shown as *P < 0.05; **P < 0.01; n.s. not significant. E. Kaplan-Meier curve showing the overall survival rate of CC patients stratified by the top 50 genes-scaled signature of Plasma/B_01_IGHA2.

Crosstalk among tumor cells, Tregs and neutrophils establishes the immunosuppressive TIME in ADC

In the above description, we have outlined that certain ADC-enriched cell sub-clusters, such as Epi_10_CYSTM1, Tregs, and pro-tumor TANs, might play oncogenic roles in regulating various cellular activities, including maintenance of stemness and evasion of immune surveillance. To gain deeper insights into the interplay among these clusters and their influence on ADC progression, we harnessed the Cellchat tool (Jin et al., 2021). Our primary focus revolved around unraveling the mechanisms governing Treg recruitment to the tumor microenvironment and the maintenance of stemness within ADC-enriched cell sub-clusters.

Given that the 3 ADC-specific sub-clusters (Epi_07_CAPS, Epi_10_CYSTM1 and Epi_12_RRAD) of epithelial cells were predicted to exhibit high levels of stem-like properties, and particularly that Epi_10_CYSTM1 presented the highest degree of malignancy among them, we further analyzed their possible interactions with other types of cells. Firstly, in comparison to SCC, ADC-enriched epithelial sub-clusters exhibited a higher propensity for interaction with Tregs through ligand-receptor signaling pathways, including ALCAM (ALCAM>>CD6) and MHC-II (HLA-DR>>CD4) (Fig. 7A-E), which have been reported to be essential for Treg recruitment, expansion and stabilization to establish the immunosuppressive microenvironment (Chalmers et al., 2022; Ferragut, Vachetta, Troncoso, Rabinovich, & Elola, 2021; Freitas et al., 2019). To further uncover the underlying reason for the enhanced stemness observed in ADC epithelial cells, we focused on the signaling received by Epi_10_CYSTM1 cell cluster. Interestingly, the Cellchat analysis revealed that, compared to other pathways, such as CD46>>JAG1 and GZMA>>PARD3, the interaction of TGFβ>>TGFBR between Tregs and epithelial cells (particularly in cluster Epi_10_CYSTM1) was activated in ADC but not in SCC (Fig. 7F-G). TGFβ is a canonical secreted protein that induces carcinogenesis, such as cancer cell proliferation, invasion, self-renewal and EMT (epithelial-to-mesenchymal transition) (Derynck, Turley, & Akhurst, 2021; Massagué, 2008; Tauriello, Sancho, & Batlle, 2022). Thus, we speculate that the recruited Tregs might secrete TGFβ to stimulate the tumor epithelial cells of ADC, leading to increased stemness. Additionally, pro-tumor TANs were also recruited to the tumor area and interacted with Epi_10_CYSTM1, as well as Epi_07_CAPS and Epi_12_RRAD cells, via the ANNEXIN-to-FPR1/FPR2 signaling pathway (Supplementary Fig. 7A-B). Previous studies have reported that the ANNEXIN pathway can promote the chemotaxis of neutrophils (Araújo et al., 2021). Likewise, TANs expressing isg, which have been shown to promote tumorigenesis as mentioned earlier (Supplementary Fig. 5C), were also recruited to the tumor area through the ANNEXIN pathway. These findings suggest that the aggressive clusters of TANs may contribute to the formation of an ADC-specific TIME.

The cellular interaction modules of sub-clusters of T cells, neutrophils and tumor epithelial cells.

A. Circle plots showing the interacting networks between epithelial cell sub-clusters and T cell sub-clusters via the pathway of ALCAM, by comparing ADC and SCC. B. Circle plots simplified from A to show the interactions among target cell clusters via ALCAM pathway. C. Circle plots showing the interacting networks between epithelial cell sub-clusters and T cell sub-clusters via the pathway of MHC-II, by comparing ADC and SCC. D. Circle plots simplified from C to show the interactions among target cell clusters via MHC-II pathway. From A to D, The direction of each arrow shows the regulation from outputting cells to incoming cells. The width of the each line shows the predicted weight and strength of regulation. E. Bubble plot showing the probability of ligand-to-receptor combination of each pathway between two different target sub-clusters of cells, by comparing ADC with SCC. F. Circle plots showing Tregs regulate epithelial cells via the TGF-β pathway, which is solely activated in ADC. G. Bubble plot showing the probability of ligand-to-receptor combination of TGF-β pathway between Tregs and epithelial cells, by comparing ADC with SCC. The pathways of ADGRE5, CD46, GZMA and NAMPT are used as negative controls. H. Dual IF staining confirming that in SLC26A3high regions of CC tissues, more FOXP3+ cells are recruited than in SLC26A3low regions (left). The numbers of recruited FOXP3+ cell are quantified using histogram plot (right). Three individual samples with ROI were calculated and P<0.01 were marked with **, showing significant difference. I. Multiplexed IF staining confirming the interaction between CD6 (on FOXP3+ cells) and ALCAM (on SLC26A3high epithelial cells) in the ALCAM pathway. J. Multiplexed IF staining showing that the recruitment of FOXP3+ cells towards SLC26A3high cells might induce EMT (marked with E-cadherin) and increase the stemness (marked with ALDH1A1) of tumor cells, via TGF-β pathway.

To test these hypotheses, we conducted immunofluorescence (IF) analysis and observed a greater recruitment of Tregs (marked by FOXP3) in close proximity to regions exhibiting high SLC26A3 expression (Epi_10_CYSTM1 cell cluster) (Fig. 7H). Furthermore, when compared to regions with low SLC26A3 expression, tumors displaying high SLC26A3 levels exhibited significantly increased expression of Aldehyde Dehydrogenase 1 Family Member A1 (ALDH1A1), a known marker for cancer stem cells in CC (Douville, Beaulieu, & Balicki, 2009) (Supplementary Fig. 7C). Notably, we identified SLC26A3high tumor cells expressing ALCAM in close proximity to CD6+ Tregs, suggesting potential involvement of ALCAM>>CD6 signaling in Treg recruitment (Fig. 7I). Additionally, we observed reduced E-cadherin expression in tumor cells within SLC26A3high regions, possibly induced by TGFβ secreted by adjacent Tregs (Fig. 7J and Supplementary Fig. 7D). Altogether, these findings confirm the intricate crosstalk between the Epi_10_CYSTM1 cell cluster and Tregs, which might help to establish an immunosuppressive TIME and sustain heightened stemness in tumor cells.

Discussion

Cervical cancer is often regarded as less aggressive compared to ovarian cancer and endometrial cancer, primarily due to the effectiveness of HPV vaccines in prevention. When diagnosed at early stages, the 5-year survival rate for cervical cancer patients exceeds 90%, surpassing that of other gynecologic malignancies (Gennigens et al., 2022; Schiffman, Castle, Jeronimo, Rodriguez, & Wacholder, 2007). However, if cervical cancer metastasizes to the intraperitoneal or retroperitoneal lymph nodes, the survival rate will decrease to 30%∼60% (Cohen et al., 2019). Worse still, the 5-year survival rate for recurrent CC patients is less than 20% (H. Li, Wu, & Cheng, 2016). Specifically, when considering the ADC type alone, the prognosis is even worse. Unfortunately, our understanding for ADC is limited. Most previous studies have concentrated on SCC type of CC, and obtained heterogenetic information derived from bulk RNA-seq method. Therefore, it is worthwhile to shift our attention towards an in-depth investigation on the disparities between ADC and SCC at the single-cell resolution.

More recently, researchers have started to utilize scRNA-seq approach to investigate the tumor microenvironment of CC. For instance, Cao et al compared 5 cases of SCC tumor tissues with paired adjacent normal tissues and described the immune landscape in which specific clusters of T and B cells exhibited immune-exhausting or activating processes (Cao et al., 2023). Another study conducted by Qiu et al aimed to elucidate the distinct molecular patterns of immune reactions between ADC and SCC in the context of different HPV infection status at the single-cell level (Qiu et al., 2023). However, due to limited samples tested in these studies, our knowledge on the TIME in CC is yet inadequate. In the current study, we have generated a genomic atlas of ADC, which may depict the landscape of TIME built by both ADC tumor cells and neighboring cells. Apart from the predominant presence of epithelial cells, the cell clusters in ADC are majorly composed of T cells, tumor-associated neutrophils (TANs), and plasma/B cells. The gene subsets identified within ADC-abundant cell sub-clusters, including Epi_10_CYSTM1, Tregs, pro-tumor TANs and P/B_01_IGHA2, exert potential roles in promoting both the carcinogenesis and immunosuppression of ADC. In the TIME of ADC, the unique module of cellular interaction has been preliminarily indicated that the immunosuppressive Tregs are recruited to tumor cells via chemo-attraction. Finally, our hypothesis from scRNA-seq data is further confirmed using clinical samples, providing evidence that SLC26A3 may be a potential biomarker for ADC (Supplementary Fig. 8).

It has been reported that immunotherapies are less effective on ADC patients, especially those who are not infected with HPV (Attademo et al., 2020; Ferrall, Lin, Roden, Hung, & Wu, 2021). As the two major components of ADC cell types in our study, T cells and epithelial cells have been implied to be the dominant regulators in the TIME due to their reciprocal interaction dynamics. We identified several ADC-specific epithelial cell clusters that may exhibit stronger stemness and aggressiveness than other clusters. Among these clusters, Epi_10_CYSTM1 seemed to have the highest level of malignancy, but the underlying mechanism behind this behavior remains unknown. The Cellchat analysis revealed that Tregs might quite actively crosstalk to Epi_10_CYSTM1. To confirm this, immunostaining for co-localization demonstrated an increased aggregation of Tregs around tumor cells exhibiting overexpression of SLC26A3. It has also been widely reported that tumor cells acquire higher stemness present more aggressive features of resistance to immunotherapy (Dianat-Moghadam et al., 2022). This study indicated that TGFβ produced by Tregs might induce the stemness of Epi_10_CYSTM1 cells by activating downstream oncogenic pathways, inducing EMT on tumor epithelial cells. Because TGFβ is a soluble protein and is hard to detect on solid tissue samples, here we only validated markers of EMT and stemness to indirectly indicate the functional existence of TGFβ, which is predicted from informatic hypothesis (Nixon, Gao, Wang, & Li, 2023). Interestingly, Epi_10_CYSTM1 cells express ALCAM, which is a membranal ligand, and bind to its receptor CD6 on Tregs, activating chemotaxis and facilitating Treg recruitment (Fig. 7I). The infiltration of a large number of Tregs in ADC may significantly accelerate immune evasion (Supplementary Fig. 8). Therefore, the presence of a higher number of Tregs infiltrating the tumor area of ADC may partially explain the increased resistance to immunotherapy.

Another interesting phenomenon observed in our sing-cell sequencing data is the involvement of TANs in the regulatory interactions within the TIME of ADC. It has been clearly investigated that pro-tumor TANs, characterized by CD66, CD11, CD170 and PD-L1, exert the oncogenic roles by promoting tumor cell proliferation, angiogenesis, inducing genetic instability, and most importantly causing immunosuppression (Jaillon et al., 2020). In this study, pro-tumor TANs are relatively enriched in ADC tissues compared to SCC, especially in HPV-negative ADC cases. This suggests that ADC-specific epithelial cell sub-clusters may actively engage in the recruitment of pro-tumor TANs.

The anti-tumor function of B cells has been identified previously (Bod et al., 2023; Helmink et al., 2020). However, our study has presented a specific cluster of plasma/B cells with tumor-promoting functions. The overexpression of gene signatures from Plasma/B_01_IGH2 cells is associated with poor regression in CC patients (Fig. 6F). Interestingly, Plasma/B_01-IGHA2 was found to be closely related to the PC Cluster (Plasma cell Cluster) in Cao’s study due to the shared gene set (Cao et al., 2023), such as IGHA1, IGHA2, IGHG2, IGHG4, etc. However, their roles were contradictory. According to that study, PC cluster was derived from the local expansion of TIME and showed potential effectiveness in anti-tumor immunity. In our case, Plasma/B_01-IGHA2 was indicated to have a reverse effect and played a tumor-promoting role in ADC. This difference in function is likely attributed to the distinct cell types and their interactions in ADC, instead of SCC. However, the underneath dynamics of B cells in ADC still needs further investigation.

Owing to post-surgical upstaging, some CC patients with stage IIIC are misdiagnosed as early-stage and have undergone radical surgeries. For these patients, the most appropriate strategy of treatment should be radiotherapy alone. However, it is challenging to avoid this issue because it is difficult to identify metastatic lymph node only through CT or MRI. To solve this problem, a reliable biomarker is required to diagnose lymph node metastasis in conjunction with radiographic tool. In this study, SLC26A3 has shown its potential as a diagnostic biomarker to predict lymph node metastasis in ADC patients. Combining two large cohorts of clinical data, we found that CC patients with ADC types were more vulnerable to be misdiagnosed for clinical staging. Derived from scRNA-seq data, SLC26A3 could be employed as a representative gene for the Epi_10_CYSTM1 cluster, which was found to be enriched in stage IIIC cases of ADC. The IHC staining further confirmed that ADC patients at stage IIIC exhibited a higher expression of SLC26A3 compared to early stages (FIGO stage I∼ II). Furthermore, we tested on biopsy samples and it implied that SLC26A3 might serve as a potential predictor of LN metastasis. Our study for the first time provides evidence that ADC patients are at a higher risk of developing lymph node metastasis, which can be challenging to detect solely via radiological imaging techniques. However, by incorporating lymph node imaging with biomarkers such as SLC26A3, it may probably reduce the misdiagnosis rate among FIGO IIIC patients.

In this study, our objective is to unravel the genomic characteristics of ADC in the hope of identifying gene signatures that can elucidate its aggressive nature. SLC26A3 is selected from the DEGs enriched in cluster Epi_10_CYSTM1, and is identified to specifically represent this ADC-specific cluster. However, based on our data, we have inevitably encountered two questions: (1) the existence of cluster Epi_10_CYSTM1; (2) the specificity of SLC26A3. As for the first question, Epi_10_CYSTM1 is distinctive from the other 11 epithelial cell clusters, with specific DEGs (Fig. 2A-C). Although Epi_10_CYSTM1 is dominantly derived from sample S7, however, its proportion is relatively high when compared with other clusters such as Epi_08_SCGB3A1, Epi_09_SST and Epi_11_REG1A (Fig 2D and Supplementary Fig. 2A-B). Moreover, we have detected a positive expression of this cluster on tissue samples, whether by using ORM1/2 or SLC26A3 antibodies (two candidate markers of cluster Epi_10_CYSTM1) via IHC. As a result, we have identified the existence of this cell cluster. As for the second question, SLC26A3 has demonstrated a higher specificity compared to other candidates, such as CYSTM1, CA9, ORM1 and ORM2, following the process of identifying markers (Fig. 3A-B and Supplementary Fig. 3E). Although the IHC results indicate that SLC26A3 is positively expressed in stage IIIC cases, however, the expression is confined to specific regions (Supplementary Fig. 9), which aligns with the scRNA-seq findings for SLC26A3. Notably, according to Table 3 and Table 4, most of the clinical cases have a lower expression of SLC26A3. Taken together, SLC26A3 is a specific representative of cluster Epi_10_CYSTM1.

In summary, we characterized the genomic landscape of the TIME in ADC at a single-cell resolution. Our study revealed the presence of specific epithelial cell clusters in ADC that exhibited more aggressive features. The enriched gene signatures from these clusters may serve as potential therapeutic targets to improve the efficacy of immunotherapy. The interactions between tumor epithelial cells and other types of cells, such as T cells and neutrophils, appropriately depict a highly immunosuppressive microenvironment of ADC. Moreover, our investigation extended to the issue of post-surgical upstaging at the single-cell level, with the identification of biomarkers from specific tumor cell clusters showing potential as diagnostic indicators for the precise diagnosis of cervical cancer patients.